---
title: "01_run_nf_core_RNAseq_pipeline"
author: "JR"
date: "8/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(tidyverse)
# library(tximport)
library(DESeq2)
library(ggrepel)
library(pheatmap)
```


# Run NF_CORE RNaseq pipeline
Let's do some RNAseq !

Goal: Align reads and get the number of reads in each gene in the genome using nf-core/rnaseq 

To use this new version of the nf-core/rnaseq pipeline (v3.4) 
So what we need is

design.csv
nextflow.config file
run.sh

# config file
Let's start with the config file -- very similar to nf-core/chipseq
```{bash config file}

process {
  executor='slurm'
  queue='short'
  memory='16 GB'
  maxForks=10
}

# go to correct working directory .../analysis/18_running_RNA_seq
nano nextflow.config

# paste in text above

```

# Run.sh file
Now the run.sh file that gives nf-core/rnaseq all the instructions for nextflow to task.
```{bash run.sh file}

# Here are all the slurm instructions:

#!/bin/bash
#SBATCH -p long
#SBATCH --job-name=HEPG2_rna_seq
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=your_emailn@colorado.edu
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --mem=6gb
#SBATCH --time=10:00:00
#SBATCH --output=nextflow.out
#SBATCH --error=nextflow.err

pwd; hostname; date
echo "Here we go You've requested $SLURM_CPUS_ON_NODE core."

module load singularity/3.1.1

# Now the inputs for nextflow.
# I resorted to older version as data does 
# not pass more recent versions -- thus
# we don't need a design file after all but you will in your own work

# first telling which pipeline to run
nextflow run nf-core/rnaseq -r 1.4.2 \

# resume is awesome and let's us pick up where we left off
-resume \

# setting up a singularity container for reproducibility 
-profile singularity \

# wher is the data :)
--reads '/scratch/Shares/rinnclass/CLASS_2023/<your_folder>/CLASS_2023/CLASSES/05_R_analyses/analysis/05_RNAseq/00_RNAseq_download_NF_core_pipeline/data/fastq/*{_read1,_read2}.fastq.gz' \

# genome file
--fasta /scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/GRCh38.p13.genome.fa \

# genome annotation file
--gtf /scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/gencode.v32.annotation.gtf \

# aligner -- unique to 1.4.2 others use different parameters.
# IMPORTANT TO CHECK ALIGNER INSTRUCTION (but will get an error if wrong :)
--pseudo_aligner salmon \

# Telling NF_CORE that we imported gencode genome annotations
--gencode \

# who to tell when job is done
--email your email@colorado.edu \
# what the config file is named (note using c language)
-c nextflow.config

# date times out
date

```

# creating run.sh
Here is the run.sh in full form for nano 
Go to /analysis/18_running_nf_core folder, nano and paste in below 

# NOTE you will need to change email and path to fastq -- use same ending path  synta
```{bash run.sh}

#!/bin/bash
#SBATCH -p long
#SBATCH --job-name=HEPG2_rna_seq
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=your_email@colorado.edu
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --mem=6gb
#SBATCH --time=20:00:00
#SBATCH --output=nextflow.out
#SBATCH --error=nextflow.err

pwd; hostname; date
echo "Here we go You've requested $SLURM_CPUS_ON_NODE core."

module load singularity/3.1.1

nextflow run nf-core/rnaseq -r 1.4.2 \
-resume \
-profile singularity \
--reads 'your_path/17_API_RNASEQ/fastq/*{_read1,_read2}.fastq.gz' \
--fasta /scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/GRCh38.p13.genome.fa \
--gtf /scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/gencode.v32.annotation.gtf \
--pseudo_aligner salmon \
--gencode \
--email your_email@colorado.edu \
-c nextflow.config

date
```

# design file 
We no longer need the design file since this data only runs on onlder pipeline.
However if you were to use the new 3.4 this would be required with following format:

# sample,replicate,fastq_1,fastq_2,strandedness

NF_Core RNAseq is now on 3.6 and this data all fails on anything 2.0 or higher.
So we don't need this but in the future it is required in NF_CORE RNAseq

We made the design file in 17 from code someone else could
run and get the exact same design file !

# Let's make the following files in analysis/18_ 

# NEXTFLOW.CONFIG

```{bash}

process {
  executor='slurm'
  queue='short'
  memory='16 GB'
  maxForks=10
}

```

# LET'S RUN THE PIPELINE!

# go to right directory
cd /scratch/Shares/rinnclass/CLASS_2022/ ..... 18_running_RNAseq_NF_CORE/NF_CORE_RUN

sbatch run.sh

tail -f nextflow.out

Voila you are doing RNAseq!


```{r}
#!/bin/bash
#SBATCH -p long
#SBATCH --job-name=2023_HEPG2_rna_seq
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=john.rinn@colorado.edu
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --mem=6gb
#SBATCH --time=20:00:00
#SBATCH --output=nextflow.out
#SBATCH --error=nextflow.err

pwd; hostname; date
echo "Here we go You've requested $SLURM_CPUS_ON_NODE core."

module load singularity/3.1.1

nextflow run nf-core/rnaseq -r 1.4.2 \
-resume \
-profile singularity \
--reads '/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/05_R_analyses/analysis/05_RNAseq/00_RNAseq_download_NF_core_pipeline/data/fastq/*{_read1,_read2}.fastq.gz' \
--fasta /scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/GRCh38.p13.genome.fa \
--gtf /scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/gencode.v32.annotation.gtf \
--pseudo_aligner salmon \
--gencode \
--email john.rinn@colorado.edu \
-c nextflow.config

date
```



