---
title: "01_binding_vs_expression"
author: "JR"
date: "8/9/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
source("../../../util/intersect_functions.R")
source("../../../util/_setup.R")
library(ggpubr)


```

# Reading in sample sheet

```{r read in sample_sheet}

# First let's read in the sample sheet to know what is what
samplesheet <- read_rds("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/19_rnaseq/final_samplesheet.rds")

```

# reading in TPM values from Salmon
# Transcripts per million 
These are different than the raw counts needed as in put for DEseq
Those are used for normalization and differential expression
Also for making heatmaps (could also do with TPM as well)
TPM and counts are used for different things but you can 
Find diffrential expressed genes and then make a gene list
Then plot TPM values of those gene list -- so they are inter-relatable

Now let's read in the TPMs from salmon
```{r reading in salmon Salmon TPMs}

# reading in salmon counts TPM (note we previously read in counts)
# TPM is transcripts per million read -- so is a normalized abundance value
salmon_tpm <- read.csv("/scratch/Shares/rinnclass/CLASS_2022/ruthquispe/CLASS_2022/class_exeRcises/analysis/18_running_RNAseq/Nf_core_run/results/salmon/salmon_merged_gene_tpm.csv")

# salmon_tpm <- read.csv("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/18_running_RNAseq_NF_CORE/preclass_NF_core_RUN/results/salmon/salmon_merged_gene_tpm.csv")


# since we have replicates we will take the mean TPM for each fractionation.
# while we are at it we will add more intuitive row and col names 
# finally we are making sure the TPM table is in same order as samplesheet

tpm <- salmon_tpm %>% 
  # pivot longer here is going to take cols 2 through the rest (the samples conditions)
  # So each gene will have 10 rows with the sample_id = experiment name/condition
  # TPM will be TPM 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_id", values_to = "tpm") %>%
  # This will merge the sample sheet by sample_id column
  merge(samplesheet) %>%
  # we group the data by gene_id condition thus there are 2 values in 5 condition
  group_by(gene_id, condition) %>%
  # taking the mean by using summarize function (this is the mean of 2 res in each condition)
  summarize(tpm = mean(tpm, na.rm = T)) %>%
  # bringing the table back to "normal" with pivot wider
  # the condition will be come the cols with names_from = with a prefix tpm_condition name
  # So we are back to 5 columns since each condition has only one mean value now
  # Those TPM mean values are populating the rows now
  pivot_wider(names_from = condition, values_from = tpm, names_prefix = "tpm_")

```

# reading in promoter peak features data.frame from 12_.Rmd
Now let's read in the promoter overlaps (peak_occurrence_df)
how many DBPs are bound at each promoter

```{r loading in peak features data frame }

# loading peak_feature
promoter_features_df <- read.csv("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/peak_occurence_dataframe.csv")

# let's take a look and refresh our memories of this data frame we created of ChiP binding properties
View(promoter_features_df)

# promoter_features_df <- read.csv("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/peak_occurence_dataframe.csv")

# Now we can merge in the TPM data to this data.frame
# This will merge on the on teh gene_id column
promoter_features_df <- merge(promoter_features_df, tpm)

# saving this file
write.csv(promoter_features_df, "promoter_feature_df_tpm.csv")

# let's plot the number of DBPs bound as a densuty plot.
ggplot(promoter_features_df, aes(x = number_of_dbp)) +
  geom_density() 

# Cool we see the bimodal distribution again.

# Let's save
  ggsave("figures/DBP_binding_density_plot.pdf")

# This is a very cool result there must be two classes of promoters!
# Remember this plot :) !
# yet I guess we already knew this from 14_2

```

RESULT: There is a bimodal distriubution of the number of DBPs bound
to a given promoter!

One that falls off around 100 DBPs and another
that emerged at ~250 DBPs on a promter. That's a lot of proteins on
one promoter!


# Abundance of genes in each cellular fraction
Plotting the abundance of genes in each cellular fraction.
```{r TPM of genes in each fraction}

# First we need to the tpm DF into a matrix
tpm_matrix <- tpm %>% 
  column_to_rownames("gene_id") %>%
  as.matrix()

# And z-scale each row.
tpm_scaled <- t(scale(t(tpm_matrix)))

# We want to be sure that we're not removing rows that have NA in only one fraction.
# And right now we're not sure.
summary(is.na(tpm_matrix))

# checking TPM scaled
summary(is.na(tpm_scaled))

# NA = false, cool, but if there were we would want to remove them with complete.cases:
?complete.cases

# clearing all possible NAs
tpm_scaled <- tpm_scaled[complete.cases(tpm_scaled),]

# Now we can hake a heat map of tpm_scaled (take ~5 min?)
pheatmap::pheatmap(tpm_scaled, show_rownames = FALSE)
# save this in the plots window >> MAKE SURE TO CHANGE WORK DIR to FIGURES

# RESULT: 
# (1) Most RNAs are abundant in the nucleus
# (2) Some RNAs expressed in total that are not in other fractioins?
```

# Plotting binding versus expression

Now let's examine how binding effects expression.
We have published previously that the more binding
events at a promoter the more abundant the expression is.
Let's see if this holds for this larger number of DBPs
```{r DBP promoter binding versus total RNA expression}

# We will plot:
# x-axis = number of DBPs bound on the promoter
# y-axis = expression level of that gene.

# y-axis total RNA TPM (+0.0001 to avoid 0 issues)
       # x-axis is number of DBPs
ggplot(promoter_features_df, 
            aes(y = log2(tpm_homo_sapiens_hepg2 + 0.001), x = number_of_dbp, color = gene_type)) + 
geom_point(data = promoter_features_df %>% filter(tpm_homo_sapiens_hepg2 < 0.001),
             shape = 17, alpha = 0.7) +
  
  # Adding a generative additive model 
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  # this adds the statistics from the gam to the figure
  stat_cor() +
  # this is just to make it look nice.
  scale_x_continuous(expand = c(0,0)) +
  # adding colors manually
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  # title
  ggtitle("Expression vs. promoter binding events") + 
  # labeling axes
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM))) 
  # saving figure
  ggsave("figures/expression-vs-promoter_binding.pdf")


# The ones with high number of DBPs and no expression are our reservoirs
```
RESULTS:
(1) Expression is typically linearly correlatd with the number of DBPs at a proomter.
This validates known data from Mele et al. on a much larger scale.

(2) Linear correlation of DBP to expression level is similar for lncRNA and mRNA.
This is also validates that lncRNAs are typically lower abundant. 

(2) There is a very interestingn line on the bottom of the plot !
This means that there is a unique distribution of promoters that
are bound by many DBPs (up to 400!).

# Binding versus nuclear expression
Let's see if the binding versus expression holds in the nuclear fraction
```{r binding versus nuclear expression}

# Now let's make a similar plot for nuclear RNA abundance versus #DBPs bound to their promoter
ggplot(promoter_features_df, 
            aes(y = log2(tpm_homo_sapiens_nuclear_fraction + 0.001), x = number_of_dbp, color = gene_type)) + 
  geom_point(data = promoter_features_df %>% filter(tpm_homo_sapiens_nuclear_fraction < 0.001),
             shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  stat_cor() +
  scale_x_continuous(expand = c(0,0)) +
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  ggtitle("Nuclear Expression vs. promoter binding events") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM))) 
  # saving figure
  ggsave("figures/nuclear_expression-vs-promoter_binding.pdf")

```
RESULT: looks very similar to total RNA binding versus expression


# Binding versus cytoplasmic expression

Next we will determine the DBP binding versus cytoplasmic expression
```{Binding versus cytoplasmic expression}

# Same thing just seeing if there is a difference of cyto RNAs versus DBPs on promoter
ggplot(promoter_features_df, 
            aes(y = log2(tpm_homo_sapiens_cytosolic_fraction + 0.001), x = number_of_dbp, color = gene_type)) + 
            # note if we take off the log2 everything is very lowly expressed
  geom_point(data = promoter_features_df %>% filter(tpm_homo_sapiens_cytosolic_fraction < 0.001),
             shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  stat_cor() +
  scale_x_continuous(expand = c(0,0)) +
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  ggtitle("Cytoplasmic Expression vs. promoter binding events") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM))) 

  # saving figure
  ggsave("figures/cytoplasmic_expression-vs-promoter_binding.pdf")
  
  # slightly fuzzier !

```
RESULTS:
(1) everything seems to be low abundance 
(2) Some mRNAs are expressed in the nucleus -- we could look at this more later.
(3) The same linear trend holds but is driven by mostly low expression events.

# lncRNA versus mRNA expression in total RNA
Next we will directly test the lncRNA vs mRNA expression levels in total RNA. 
```{r determining lncRNA and mRNA expression levels in total RNA}

# We can see this from the trend lines that lncRNAs are less expressed,
# but here's another way of looking at that.

# the main function here is color by gene_type in promoter_features_df
ggplot(promoter_features_df, aes(x = log2(tpm_homo_sapiens_hepg2 + 0.01), color = gene_type))+
  geom_density()

# saving figure
ggsave("figures/mrna_lncrna_tpm_total_rna.pdf")

# let's also do the same for nuclear since lncRNAs are typically more nuclear
ggplot(promoter_features_df, aes(x = log2(tpm_homo_sapiens_nuclear_fraction + 0.01), color = gene_type))+
  geom_density()

# saving figure
ggsave("figures/mrna_lncrna_tpm_nuclear.pdf")


```
RESULT:
(1) This yet again confirms lncRNAs have lower expression levels than mRNAs.
(2) In the nuclear fraction it shift's to closer.
(3) lot's of mRNA with nuclear expression -- that seems odd



We have previously observed that k562 cells also exhibit high binding promoters
that are not expressed. We termed them 'reservoirs' as they are a reservoir
for many Dna-protein interaction sites. Based on the plot above we observed that
this phenomena also exists in hepG2 cells as well. 

Based on this we next wanted to identify the 'reservoirs' in hepG2 cells.
```{r defining HEPG2 reservoirs}

# first we will use a cutoff of 100 DBPs.
# We can think about this more in the future.

promoter_features_df$hepg2_reservoir <- 
  as.numeric(promoter_features_df$number_of_dbp > 100 & 
               promoter_features_df$tpm_homo_sapiens_hepg2 < 0.001)

# seeing what we got with table
table(promoter_features_df$hepg2_reservoir)

# RESULT:
# (1) There are 798 reservoirs in HEPG2.

```

Now that we have defined reservoirs in hepG2 cells, we next want to determine how many 
are similar genomic regions in k562 and hepG2.

```{r reading in K%62 reservoirs}

k562_df <- read_csv("/scratch/Shares/rinnclass/CLASS_2022/data/2020_k562_promoter_peak_df.csv")

# saving for future use this was hard to find :)
write_csv(k562_df, "2020_k562_promoter_peak_df.csv")

# next we want to merge the k562 adn Hepg2 DFs 
# first we should label what is k562 and what is Hepg2


# K562_df renaming :

k562_df <- k562_df %>% 
  dplyr::select(gene_id, reservoir, conservative_reservoir, tpm, expression, tf_binding, promoter_mean_tpm, promoter_median_tpm, promoter_max_tpm) %>%
  dplyr::rename(k562_reservoir = reservoir, 
                k562_conservative_reservoir = conservative_reservoir,
                k562_expression = expression,
                k562_tpm = tpm,
                k562_tf_binding = tf_binding,
                k562_promoter_mean_tpm =  promoter_mean_tpm,
                k562_promoter_median_tpm = promoter_median_tpm,
                k562_promoter_median_tpm = promoter_median_tpm,
                k562_promoter_max_tpm = promoter_max_tpm)

# save this file in new format
write_csv(k562_df,"k562_df.csv")

# testing read in for future use
k562_df <- read_csv("k562_df.csv")

# renaming promoter_features_df to hepg2_df

hepg2_df <- promoter_features_df %>%
  dplyr::select(gene_id, gene_name, tpm_homo_sapiens_hepg2, tpm_homo_sapiens_cytosolic_fraction, tpm_homo_sapiens_nuclear_fraction, tpm_homo_sapiens_insoluble_cytoplasmic_fraction, tpm_homo_sapiens_membrane_fraction, number_of_dbp, hepg2_reservoir) %>%
   dplyr::rename( tpm_total = tpm_homo_sapiens_hepg2,
                 tpm_cytosolic_fraction =  tpm_homo_sapiens_cytosolic_fraction,
                 tpm_nuclear_fraction = tpm_homo_sapiens_nuclear_fraction ,
                 tpm_insoluble_cytoplasmic_fraction = tpm_homo_sapiens_insoluble_cytoplasmic_fraction ,
                 tpm_membrane_fraction = tpm_homo_sapiens_membrane_fraction)

# let's save this handy file
write_csv(hepg2_df,"hepg2_df.csv")

# testing read in for future
hepg2_df <- read_csv("hepg2_df.csv")
  
# Let's merge the k562 reservoirs in with HEPG2_df
# Merges on Gene_id
hepg2_k562_promoter_features_df <- merge(hepg2_df, k562_df)

# Now saving
write_csv(hepg2_k562_promoter_features_df, "hepg2_k562_promoter_features_df.csv")

# Make a table of reservoir status
res_status <- hepg2_k562_promoter_features_df %>% 
  #  !! another useful combination of group_by and summarize !!
  group_by(hepg2_reservoir, k562_reservoir, k562_conservative_reservoir) %>%
  summarize(count = n())

# saving for future
write_csv2(res_status, "reservoir_overlap_stats.csv")
```

# Writting out files

```{r saving files for future use}

# We can now write these out for safekeeping / use in other analyses
write_csv(promoter_features_df, "results/promoter_features_df.csv")
write_csv(tpm, "results/mean_tpm_per_condition.csv")
write_csv(samplesheet, "results/samplesheet.csv")
```

There's a lot to explore here. We have some potential analyses that you can pursue, or your group can decide on a direction that you're interested in.

1. How does the binding vs. expression change for the high binding promoters (> ~300 DBPs) vs low binders? And does this trend change across subcellular fractions?
2. For the genes that are exclusively in one fraction (i.e. exclusively nuclear) what do their promoter binding profiles have in common?
3. Are there any mRNAs that are exclusively in the nuclear fraction?
4. What does the expression of the K562 reservoirs look like in HEPG2 in terms of expression? Are they more lowly expressed? Or does their trend follow the binding vs expression trend for non-K562 reservoirs?