---
title: "01_test_knit"
author: "JR"
date: "3/17/2022"
output: github_document 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(GenomicRanges)
library(tidyverse)
library(IRanges)
source("../../../util/my_class_functions.R")

# filepath to import peaks
basepath <- "/scratch/Shares/rinnclass/CLASS_2023/JR"
peak_path <- "CLASS_2023/CLASSES/03_Nextflow/00_my_chipseq/results/bwa/mergedLibrary/macs/broadPeak/"
broadpeakfilepath <- file.path(basepath, peak_path)
```
Acompanying youtube video:
https://youtu.be/BPHQAMAh5GE

Goal: to create a consensus_peak file for each dbp, export and format for UCSC.
The first step in analyses is looking at raw data :)
Here we will take our functions to create a consensus peak file and see how it 
lines up with raw data on the UCSC genome browser. We will do this in these steps:

1) Import peaks
2) create consensus peaks
3) export named GRanges list to .bed files
4) format the file to be uploaded into UCSC
5) inspect raw data to see if consensus peaks make sense.

# STEP1: import peaks 
```{r import peaks}

# import peaks
peak_list <- import_peaks(consensus_file_path = broadpeakfilepath)

# let's get a list of how many peaks are in each file before we create consensus peaks.
peak_num <- sapply(peak_list, length) %>% 
  as.data.frame(row.names = T)
# label column
names(peak_num) <- c("num_peaks")

# make dbp name a col.

peak_num <- peak_num %>%
  rownames_to_column(var = "dbp") %>%
  
  # This is pretty similar to text to columns in excel
  separate(col = dbp,  into = c('dbp', 'replicate'), sep = "_")
  # peak_num <- separate(peak_num, col = dbp,  into = c('dbp', 'replicate'), sep = "_")
peak_num