---
title: "02_DESEQ_differential_expression"
author: "JR"
date: "8/9/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(tidyverse)
library(pheatmap)
library(DESeq2)
#install.packages("ggrepel")
library(ggrepel)
source("/scratch/Shares/rinnclass/CLASS_2023/Christopher/CLASS_2023/util/plotting_functions.R")
source("/scratch/Shares/rinnclass/CLASS_2023/Christopher/CLASS_2023/util/_setup.R")

```
Accompanying YouTube Video
https://youtu.be/fwcuRmVWpOM


# Differential Expression (nuclear versus cytoplasmic)
In the last class we created normalized counts that are ready to compared across
conditions using DESEQ. WE can factor any sample in our samplesheet to compare
to the other conditions. DESEQ will then calculate which genes are significantly 
differentially expressed between the factored samples.

Note we "factored" the sample sheet so each fraction has
a 'factor' value. So we can now factor nuclear and compare to cytoplasmic.

# To learn more about DeSeq and other variations not used here:
https://physiology.med.cornell.edu/faculty/skrabanek/lab/angsd/lecture_notes/08_practical.pdf

GOAL: today we will compare nuclear and cytoplasmic expression to find those
genes that are specific to nuclear or cytoplasmic fractions of the cell.

# Loading in the enviroment objects from previous class on creating counts
```{r laoding counts from DESEQ and input counts and files needed (e.g., samplesheet)}

# erase environment and reload!
load("results/DESEQ_RLOG.RData", verbose = T)

```

Let's specialize sample sheet we can use to compare nuc/cyto samples.
##### FACTORING IS CRITICAL #####
```{R nuclear and cytoplasmic samplesheet}

# Now we will filter the sample sheet to just nuclear and cytoplasmic
samplesheet_nuc_cyto <- samplesheet %>%
  filter(condition %in% c("homo_sapiens_cytosolic_fraction", "homo_sapiens_nuclear_fraction"))
# Nice now we have a reduced sample sheet that is well organized.

# Next, let's triple check that the counts data
# contains the counts for the samples in the reduced sample sheet 
samplesheet_nuc_cyto$sample_name %in% colnames(counts)
# Cool, the 4 samples in filtered are same as counts

# Again we need to think about "rownames" we will convert 
# samplesheet_nuc_cyto to df and add rownames by indexing
samplesheet_nuc_cyto <- samplesheet_nuc_cyto %>%
  as.data.frame()

# adding row names
rownames(samplesheet_nuc_cyto) <- samplesheet_nuc_cyto$sample_name

# NOTE: DeSeq wants the control as a factor (fct), currently (chr)
# The first factor level is the "control" and all the other levels
# will be compared back to it.

# Let's make the condition a factor:
samplesheet_nuc_cyto$condition <- factor(samplesheet_nuc_cyto$condition, 
                                         levels = c("homo_sapiens_nuclear_fraction",
                                                    "homo_sapiens_cytosolic_fraction"))

# IMPORTANT : values returned will be nuclear/cytoplasmic !!
# IMPORTANT : !! FIRST FACTOR is denominator !!


# let's write out this samplesheet for future use
write_rds(samplesheet_nuc_cyto, "results/samplesheet_nuc_cyto.rds")

```


# Heatmap of all samples (Rlog Counts)
First let's cluster all the data to see how the samples relate.
```{r heatmap of all data}

# Now we can make a heatmap of all the different fractions
# First let's look at the rlogCounts
hist(rlog_counts_matrix, breaks = 100, xlim = c(0,20))

# First we will filter to genes that are variable across data.
# rowVars gives us the variance across samples
hist(rowVars(rlog_counts_matrix), breaks = 100, xlim = c(0,4))

# let's look at this as an object
x <- rowVars(rlog_counts_matrix)
hist(x)
# Filtering out genes that don't have variance of more than 1 log2 of variance
rlog_var_genes_all <- rlog_counts_matrix[rowVars(rlog_counts_matrix) > 1,]

# row center the counts -- we need to flip the matrix
# we can only scale cols so we need to transmute (t)
# then turn it back with (t)
scaled_counts <- t(scale(t(rlog_var_genes_all))) %>%
  as.matrix()

# make heat map
pheatmap(scaled_counts, show_rownames = FALSE)


#TODO r set up file figure colors
# saving
pdf("figures/all_vs_total_genes_heatmap.pdf")
pheatmap(scaled_counts, show_rownames = FALSE)
dev.off()
```

# Principal component analysis (Rlog Counts)
Now let's use Principle Component Analysis to determine sample similarity

```{r PCA analysis}

# The first thing that we'll want to do is make a PCA plot.
# 'prcomp()' is the principle component analysis function.
?prcomp

# First we need to flip the counts matrix as PCA needs to be across rows (samples)
rld_pca <- prcomp(t(rlog_counts_matrix))

# setting up a data frame with column x 
rld_prcomps <- rld_pca$x %>% as.data.frame() %>%
  # note that x is in rld_pca as the PCA values
  rownames_to_column("sample_name") %>%
  # selecting data we want to merge with samplesheet
  select(sample_name, PC1, PC2)
  # merging PC1 and PC2 with sample sheet via sample name
rld_prcomps <- merge(samplesheet, rld_prcomps)


# Let's make a PCA plot
ggplot(rld_prcomps, aes(x = PC1, y = PC2, color = condition, label = replicate)) +
geom_point() +
geom_text_repel()
# RESULT: PCA plot looks like replicates are more similar than fraction type!
# we are already aware of this and can relabel.

# Let's print this out
ggsave("figures/PCA_plot.pdf")

```

# Nuclear versus cytoplasmic differential expression (Salmon Counts input)
Next we will reduce out counts_matrix to just the values of nuc/cyto

```{r nuc/cyto  counts}

# first reduce the counts matrix to just nuc and cyto samples.
counts_nuc_cyto <- counts[ ,samplesheet_nuc_cyto$sample_name]

# This is a good way to triple check everything is in order.
rownames(samplesheet_nuc_cyto) == colnames(counts_nuc_cyto)
# Great this evaluated to true and thus moved on -- we would get an error message otherwise.

```

# Finalinzing samplesheet for nuc vs cyto DEseq2
We are almost ready, last things to check:

(1) DeSeq2 requires integer data from the counts

2) Deseq2 needs the condition being compared to as a "factor" 
-- this is critical --

```{R organizing nuc/cyto counts and samplesheet for deseq}

# Ok now we need to turn the counts into integer data (int)
# The first step would be to round the counts, but we have already done that.
# But good to remember to round before setting character to int.

# mode can convert chr to int, let's change everything from chr to int
mode(counts_nuc_cyto) <- "integer"

# Let's fileter to genes that have at least one count across cols.
nuc_cyto_counts_filtered <- counts_nuc_cyto[rowSums(counts_nuc_cyto) > 1,]

```

# Differential expression of nuc -vs- cyto DEseq2 (Counts)
Now let's look deeper into the results held in dds object

```{r DEseq2 nuc versus cyto}

# IMPORTANT: always double check organization of input files to DESeq!
# If this is not true the results will be a mix of wrong samples!
# We did this above but double checking here
colnames(nuc_cyto_counts_filtered) == rownames(samplesheet_nuc_cyto)

# looks good, now let's make the required DEseq raw counts
dds <- DESeqDataSetFromMatrix(countData = nuc_cyto_counts_filtered,
                              colData = samplesheet_nuc_cyto,
                              design = ~ condition)


dds <- DESeq(dds)

# IMPORTANT : we need the results name for exporting results to data.frame
# We can see that DeSeq makes it's own names so "resultNames" is the way to get that info
resultsNames(dds)
# We don't want the intercept data
# but we need to extract the results from dds:
# let's remove intercept:

res <- results(dds, name = "condition_homo_sapiens_cytosolic_fraction_vs_homo_sapiens_nuclear_fraction")
# It's easier to view it as a data.frame so we'll convert it.
# While we are at it we merge in gene_symbols (g2s)

res_df <- res %>% as.data.frame() %>%
  # moving the row values to a "meta" column
  rownames_to_column("gene_id") %>%
  # merging in gene_name with gene_id
  merge(g2s)

# Nice now we have a DF with a lot of good info for each gene.
# Let's write it out.

write_rds(res_df, "results/nuclear_cyto_results_df.rds")

```

QC plot: Volcano 
```{r valono, MA and other plots}
# Test is CYTO/Nuclear
# so larger values are cyto and smaller nuclear

# A first check is a "volcano" plot:
# x-axis = log fold change
# y-axis = pval
# Plotting volcano plot of nuclear vs cytoplasmic

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = -log10(res_df$padj) > 90)) + 
  geom_point() 

# saving volcano plot
ggsave("figures/nuc_cyto_volcano.pdf")
```

We can see several highly cyto and nuclear genes
Let's find out who they are !

```{r inspecting valcano plot results}

# Note there was a gene log2 -30 fold down meaning very nuclear
nuclear_gene <- res_df$gene_name[which.min(res_df$log2FoldChange)]
nuclear_gene
# Nice U2 is very nuclear and good control

# we can find most cytoplasmic here too
cyto_gene <- res_df$gene_name[which.max(res_df$log2FoldChange)]
cyto_gene
# 7SK -- a lncRNA ...


# Making a list of nuclear genes at certain threshold
nuclear_genes <- res_df %>% 
  filter(log2FoldChange < -1, padj < 0.05) %>%
  as.data.frame()

hist(res_df$padj < 0.05)


# saving file
write_csv(nuclear_genes, "results/nuclear_genes.csv")

# Cytoplasmic genes:
cyto_genes <- res_df %>%
  filter(log2FoldChange > 1, padj < 0.05)

# saving file
write_csv(cyto_genes, "cyto_genes.csv")


```

# MA-PLOTS are one of the common first plots to make.
# This plot informs you of low abundance biases in fold-change

```{R MA PLOT}

# The 'MA plot' shows how low abundance affects can be an issue.
# x-axis = log(expression level)
# y-axis = log fold change
ggplot(res_df, aes(x = log10(baseMean), y = log2FoldChange)) + 
  geom_point()
# This is a good QC plot and shows the fold changes
# are not biased to low "base-mean" levels

# saving MA PLOT
ggsave("figures/MA_plot_nuc_cyto.pdf")


# Now let's look at nuclear MA plot seperate:
ggplot(nuclear_genes, aes(x = baseMean , y = log2FoldChange)) + 
  xlim(0, 50000) +
  geom_point()
# hmm most have the biggest fold change are low abundance.

ggplot(cyto_genes, aes(x = baseMean , y = log2FoldChange)) + 
  xlim(0, 50000) +
  geom_point()

# here we more genes with higher abundance being significant

```

let's see the range of pvalues in nuc-vs-cyto:
```{R pvalues nuc-vs-cyto / FDR example}

# let's look at the distribution of pvalues
hist(res_df$padj)

```


# let's save our key environment variables 

```{r}

save(counts, counts_filtered, rlog, g2s, gencode_genes, rlog_counts, rlog_counts_matrix, samplesheet, file = "results/DESEQ_RLOG.RData")

```

```{R}
# WHY DOESN"T THIS WORK? Why is that a good thing?
nuc_cyto_sig_genes <- merge(nuclear_genes, cyto_genes)


# alternative
combined_nuc_cyto_sig <- bind_rows(nuclear_genes, cyto_genes)


# counts only in sig nuc cyto genes
nuc_cyto_counts_filtered <-counts_filtered[rownames(counts_filtered) %in% combined_nuc_cyto_sig$gene_id , ]
table(is.na(nuc_cyto_counts_filtered))
# scaling for a heatmap
nuc_cyto_sig_scaled_counts <- t(scale(t(nuc_cyto_counts_filtered))) %>%
  as.matrix()

table(is.na(nuc_cyto_counts_filtered))
# make heat map
pheatmap(nuc_cyto_sig_scaled_counts, show_rownames = FALSE)
#reducing to just nuc and cyto counts
just_nuc_cyto_counts_sig <-  nuc_cyto_sig_scaled_counts[, samplesheet_nuc_cyto$sample_name]
# make heatmap
pheatmap(just_nuc_cyto_counts_sig, show_rownames = FALSE)
```

# More stringent log2 FC
```{r}
# filtering more on fold change
stringent_cyto_nuclear <- combined_nuc_cyto_sig %>%
  filter(log2FoldChange > 2, padj < 0.05) %>%
  as.data.frame()


# getting counts
stringent_nuc_cyto_counts_filtered <-counts_filtered[rownames(counts_filtered) %in% stringent_cyto_nuclear$gene_id , ]


# reducing to just nuclear cyto
stringent_just_nuc_cyto_counts_sig <-  stringent_nuc_cyto_sig_scaled_counts[, samplesheet_nuc_cyto$sample_name]

#scaling
stringent_nuc_cyto_sig_scaled_counts <- t(scale(t(stringent_just_nuc_cyto_counts_sig))) %>%
  as.matrix()


# make heat map
pheatmap(stringent_nuc_cyto_sig_scaled_counts, show_rownames = FALSE)
```


```{r}
save(counts, counts_filtered, rlog, g2s, gencode_genes, rlog_counts, rlog_counts_matrix, samplesheet, file = "results/DESEQ_RLOG.RData")
```

