---
title: "01_peak_features"
author: "JR"
date: "8/3/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
source("../../../util/my_class_functions.R")

# filepath to import peaks
basepath <- "/scratch/Shares/rinnclass/CLASS_2023/JR"
peak_path <- "CLASS_2023/CLASSES/05_R_analyses/00_consensus_peaks/consensus_peaks"
consensusPeakPath <- file.path(basepath, peak_path)
```

Goal: Here we are going to create a bunch of human genome annotation objects
and how many peaks (consensus_peaks) over lap them. 

We are now going to start to create results on binding versus genome features!
More importantly we are going to save our environment at the end for quick
loading moving forward.

In short we are on our way to analyzing our custom ChIPseq data !!


# Exploring the genome features bound by DBPs
Now that we have our consensus_peaks across replicates
we can start to explore where they are bound in the genome.
Do they bind at promoters? How much of the genome do they cover?
And many more questions.

# set up working directory: 
set working directory to CLASS_2023/CLASSES/05_R_analyses/01_peak_features


# Let's start importing all the files we need:

```{r importing files}

consensus_peaks_files <- list.files(consensusPeakPath, 
                                             pattern = "*.bed",
                                             full.names = TRUE)

# lapply with import function to make a list of GRanges
consensus_peaks <- lapply(consensus_peaks_files, rtracklayer::import)

# adding DBP name to each list of GRanges from the file name
# note gsub requires no space around or argument ( | )
names(consensus_peaks) <- gsub("/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/05_R_analyses/00_consensus_peaks/consensus_peaks/|_consensus.bed","", consensus_peaks_files)

```

# Loading in Gencode annotation of genomic features
We will make a file of lncRNA and mRNA promoters from 
the larger list of GRanges in Gencode

```{R loading gencode and making lncRNA and mRNA annotation files } 

# let's start making lncRNA and mRNA annotation files that we will use moving forward 
# good way to refamiliarize with GENCODE GRanges -- The whole genome is in there :)

gencode_gr <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/gencode.v32.annotation.gtf")

# now just type == gene annotations
gencode_genes <- gencode_gr[gencode_gr$type == "gene"] 
table(gencode_gr$type)

# exporting all genes file (we will save all the .Rdata too at the end)
rtracklayer::export(gencode_genes, "results/gene_annotations/gencode_genes.gtf")

# mRNA genes (called "protein_coding") in this version of gencode changes sometimes !
mrna_genes <- gencode_genes[gencode_genes$gene_type %in% "protein_coding"] 

rtracklayer::export(mrna_genes, "results/gene_annotations/mrna_genes.gtf")
table(gencode_genes$gene_type)

# now doing a second index for lncRNA:
lncrna_genes <- gencode_genes[gencode_genes$gene_type %in% "lncRNA"] 

rtracklayer::export(lncrna_genes, "results/gene_annotations/lncrna_genes.gtf")

# both mRNA and lncRNA annotations together.
mrna_lncrna_genes <- gencode_genes[gencode_genes$gene_type %in% c("protein_coding","lncRNA")]
rtracklayer::export(mrna_lncrna_genes, "results/gene_annotations/mrna_lncrna_genes.gtf")

```
We could do the above for thousands of annotations. But we should consider 
how big gencode_gr is. A common solution is to make GRange of needed annotations.
This loads in much faster and can be made into readable dataframe pretty easily.

# let's see what this will be like moving forward !

```{R GRanges Genes to DF} 
# starting annotation file that we will use moving forward.
lncrna_mrna_genes <- rtracklayer::import("results/gene_annotations/mrna_lncrna_genes.gtf")

# Nice that loaded so much faster -- lets see as DF
lncrna_mrna_genes_df <- lncrna_mrna_genes %>% as.data.frame()
# easy to read and tons of info !
```

Now let's make a promoter annotation file and gene_body for future use.
```{R promoter annotations} 
# promoter regions with Genomic Ranges funciton "promoters"
?promoters

# creating and exporting promoter annotations
lncrna_mrna_promoters <- promoters(lncrna_mrna_genes, upstream = 1000, downstream = 1000)
# check right size
width(lncrna_mrna_promoters)
rtracklayer::export(lncrna_mrna_promoters, "results/gene_annotations/lncrna_mrna_promoters.gtf")


# last handy annotation will be lncRNA and mRNA gene IDs to subset
lncrna_gene_ids <- mrna_lncrna_genes$gene_id[mrna_lncrna_genes$gene_type == "lncRNA"]
table(mrna_lncrna_genes$gene_type)
# same for mRNAs
mrna_gene_ids <-mrna_lncrna_genes$gene_id[mrna_lncrna_genes$gene_type == "protein_coding"]

# note we could have indexed into promoters too -- just getting gene_id
```

Now that we have all the annotations we need let's start seeing
what annotations overlap the peaks in our consensus peaks!

# How many peaks overlap gene annotations ??
Let's analyze how many peaks overlap different genomic features
we can start by creating a data frame of peak overlaps.
```{r creating an overlap features data.frame}

# Let's start with loading in the number of peaks each DBP has -- using length.

num_peaks_df <- data.frame("dbp" = names(consensus_peaks),
                           "num_peaks" = sapply(consensus_peaks, length))


# Now let's get the total amount of the genome covered by all the peaks for a given DBP.

num_peaks_df$total_peak_length <- sapply(consensus_peaks, function(x) sum(width(x)))
# nice we are getting results !

```

# expanding our num_peaks_df information
Great we are starting to build an informative DF to incorporate our results.
Now let's start looking at how many peaks overlap the genome feature files we made.

# count_peaks_per_feature function to find overlaps
we made a handy function that basically does
"countOverlaps" of number of times a feature overlaps.
Take a look at the function -- pretty simple.
@parameters:
(1) annotation file in GRanges
(2) peak file
(3) counts (number of overlaps), 
occurence (matrix of 1 if overlap 0 if not)

```{r count_peaks_per_feature function !}

# counting promoter overlaps
promoter_peak_counts <- count_peaks_per_feature(lncrna_mrna_promoters, consensus_peaks, type = "counts")

# ok we see this is a large matrix of the number of overlaps at each promoter
# cols are promoters, rows are DBPS
# so now we just row sum for each DBP !

num_peaks_df$peaks_overlapping_promoters <- rowSums(promoter_peak_counts)

# Now let's break these promoters into two groups "lncrna" and "mrna"
# We will use the gene_id objects we made above to index and separate them.

num_peaks_df$peaks_overlapping_lncrna_promoters <- rowSums(promoter_peak_counts[,lncrna_gene_ids])

# mrna promoter overlaps
num_peaks_df$peaks_overlapping_mrna_promoters <- rowSums(promoter_peak_counts[,mrna_gene_ids])

# Let's take a look.
write_csv(num_peaks_df, "results/num_peaks_df.csv")

```



# extracting gene_body_peak_counts data 
Now let's do the same thing for gene-bodies
First we will make a gene-body object like we did for promoters above.

```{r extracting genebody overlpas counts data }

# Finding overlaps with gene_bodies (will take a few minutes again)
# Note this takes 10-15 min
genebody_peak_counts <- count_peaks_per_feature(mrna_lncrna_genes, 
                                                consensus_peaks, 
                                                type = "counts")

# Now let's extract the overlaps the same way we did for promoters above

# All gene bodies
num_peaks_df$peaks_overlapping_genebody <- 
  rowSums(genebody_peak_counts)

# lncRNA gene bodies 
num_peaks_df$peaks_overlapping_lncrna_genebody <- rowSums(genebody_peak_counts[,lncrna_gene_ids])

# mRNA gene bodies
num_peaks_df$peaks_overlapping_mrna_genebody <- 
  rowSums(genebody_peak_counts[,mrna_gene_ids])

# let's take a look.
# cool lots of info let's save:
write_csv(num_peaks_df, "num_peaks_df.csv")

```

# Downloading TF annotation data set
# this is a handy lesson of how to download any file in R
We have a large group of different DBPs so let's import some annotaitons of which
ones are transcription factors etc. A paper in Cell is one of the better annotations --
however they are never perfect :)

Human TF annotations:
https://www.cell.com/cms/10.1016/j.cell.2018.01.029/attachment/ede37821-fd6f-41b7-9a0e-9d5410855ae6/mmc2.xlsx

Let's download the file using R (we could also use curl/wget in BASH)
```{r downloading TF annotation data}

url <- "https://www.cell.com/cms/10.1016/j.cell.2018.01.029/attachment/ede37821-fd6f-41b7-9a0e-9d5410855ae6/mmc2.xlsx"

destination_for_url <- "results/TF_annotations.xlsx"

# to download we can use download.file
download.file(url, destination_for_url)

# The above is a handy bit of code for downloading data into R directly,but BASH works too

```

# reading in DBP annotations as transcription factor
Now that we have the file we need to read it in. Since it's excel file we will use
a new 'read' function (we have used 'read.table' & 'read.csv' etc).

To this end we will use readX1 'read_excel' function.
```{R merging in TF annotation data}

#redx1::read_excel to import
human_tfs <- readxl::read_excel("results/TF_annotations.xlsx",
                                sheet = 2, skip = 1)


# let's rename the 4th column to indicate if it is a TF.
names(human_tfs)[4] <- "is_tf"

# now let's intersect gene names that are in our ChIP data and has TF identity.
length(which(tolower(num_peaks_df$dbp) %in% tolower(human_tfs$Name)))
```

# merging annotation file to num_peaks_df

```{R merging annotation file to num_peaks_df}

# first let's filter and grab the first 4 columns that match DBPs in num_peaks_df
human_tfs <- human_tfs[tolower(human_tfs$Name) %in% tolower(num_peaks_df$dbp), 1:4]


# adding new column names
names(human_tfs) <- c("ensembl_id",
                      "dbp",
                      "dbd",
                      "tf")

# Now comes the actual merge. Let's look at it first
# ?merge

num_peaks_df <- merge(num_peaks_df, human_tfs, all.x = T)

# Let's check how many NAs -- we should have some missing values.
dim(num_peaks_df[is.na(num_peaks_df$tf),])

# Ok looks good let's write it out:
num_peaks_df <- num_peaks_df[,1:12]
write_csv(num_peaks_df, "results/num_peaks_df.csv")

```

# Occurence matrix of overlaps
Now let's also use our 'count_peaks_per_feature' function to use the 'occurence' parameter
We will make a matrix that records all the overlaps with all promoters.
This can be used for meta plots or slecting a given gene for an overlap etc.

```{r count_peaks_per_feature occurence matrix}

# note similar as above but using type = occurnece
promoter_peak_occurence <- count_peaks_per_feature(lncrna_mrna_promoters, consensus_peaks, 
                                               type = "occurrence")

# Let's double check that all lncrna & mrna genes are accounted for:
stopifnot(all(colnames(promoter_peak_occurence) == lncrna_mrna_promoters$gene_id))

# Great we will use this quite a bit moving forward so let's write it out! 
write.table(promoter_peak_occurence, "results/lncrna_mrna_promoter_peak_occurence_matrix.tsv")

```

# Peak occurence as data.frame
Now let's make the peak occurence matrix into a DF for easier access -- but not computation.
```{R turnign peak_occurence matrix into data.frame}

# First make sure promoter_peak_occurrence and lncrna_mrna_promoters are in the same order
stopifnot(all(colnames(promoter_peak_occurence) == lncrna_mrna_promoters$gene_id))


# Now let's use the 'data.frame()' fucntion. Set up a bunch of colnames and populate them.
peak_occurence_df <- data.frame("gene_id" = colnames(promoter_peak_occurence),
                                "gene_name" = lncrna_mrna_promoters$gene_name,
                                "gene_type" = lncrna_mrna_promoters$gene_type,
                                "chr" = lncrna_mrna_promoters@seqnames,   
                                "1kb_up_tss_start" = lncrna_mrna_promoters@ranges@start,
                                "strand" = lncrna_mrna_promoters@strand,
                                "number_of_dbp" = colSums(promoter_peak_occurence))

# let's look at what we have !
# nice lots of info for 30K genes
# "counts" just gave us value for the dbp of interest


# Let's write out this data frame -- all this code & run time is now encapsulated in one .csv :)
write_csv(peak_occurence_df, "results/peak_occurence_dataframe.csv")

```

Since we will use many of these files again we can save 
the key elements as .Rdata file to loaded in next time !
```{r saving key objects in environment to load next session}

save(consensus_peaks, gencode_genes, lncrna_gene_ids, mrna_gene_ids, num_peaks_df, peak_occurence_df, lncrna_mrna_promoters, mrna_lncrna_genes, file = "results/peak_features.RData")

# awesome - now we never have to load in consensus peaks again :)
# erase environment and reload!

load("results/peak_features.RData", verbose = T)
# WOW that is super handy !!

```


#############
ExeRcise
#############

(1) use count_peaks_per_feature for a exons annotation in gencode 
-- then try adn add to num_peaks_df and save .csv


(2) Take a look at the peak_occurence_df and num_peaks_df -- find 3 things new that we haven't done yet. :

a) Which DBP has the most coverage of peaks in lncRNAs 
b) Which non TF has the most peaks and genome coverage. How does it compare to those that are not TFs
c) make a new col of percentage of peaks that overlap promoters versus gene bodies

```{r solution}
promoter_peak_counts <- count_peaks_per_feature(lncrna_mrna_promoters, consensus_peaks, type = "counts")

# getting exon features 
exons <- gencode_gr[gencode_gr$type == "exon"]

# seeing how many in here:
length(exons)
# 1372308

# using count_peaks_per_feature function to find overlaps
exon_ov <- count_peaks_per_feature(exons, consensus_peaks, type = "counts")

# merging into num_peak_df
num_peaks_df$exon_ov <- 
  rowSums(exon_ov)

# solution 2:

# percentage peaks overlapping promoters
num_peaks_df <- num_peaks_df %>%
  mutate(percent_promoter_overlap = peaks_overlapping_promoters/num_peaks )

# percentage of peaks overlapping gene_bodies
num_peaks_df <- num_peaks_df %>%
  mutate(percent_geneBody_overlap = peaks_overlapping_genebody /num_peaks )


```

