---
title: "Class_exercise"
author: "Your Name"
date: "3/16/2023"
output: html_document
---


# Load the libraries you need
# Load functions you need "my_class_functions"
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```


# load in your peak files for each replicate of each protein
# Here I am starting to analyze my data for my proteins of interest:
# proteinX, Y, Z .....
# First I will read in each replicate file
```{r load in peak files}



# printing out a table of the number of peaks in each file:
objectname

```


# Now I am going to create consensus peaks for each protein
```{r consensus peaks}


# export consensus peaks to results folder

```

# Now I am going to make my consensus peaks compatable with UCSC genome browser
```{r}


# print out consensus peak files in a results/UCSC directory

```

# I am curious if my proteins are transcription factors so I will use the annotations
# in a cell paper I found and see

```{r}

# if you leave the object name you just created in the environment
# it will print out in the knit. For example :
objectname
```




# Now I want to compare a protein with a previous analysis 
```{r}

# goto UCSC genome browser and load in a peak file for a given protein
# load in the data for the same protein from the previous analysis
# compare how your consensus peaks are similar or different to previous analyses



```


# Now I am going to determine how my peaks for each protein overlap annotations of the genome
# First I will find the overlaps between my consensus peaks with promoters of lncRNA and mRNA promoters

```{r}


# find overlaps of promoters for each protein

```

## results: 
#1) What can you determine from these overlaps?



# Now I want to compare the overlaps with lncRNA and mRNA promoters seperately 
```{r}

```
## results:
# 1) What is the difference in overlaps between mRNA and lncRNA promoters


# Now I am going to test if there is more binding over gene bodies than promoters
# I will seperate lncRNA and mRNA gene bodies to find the overlaps 

```{r}



```
## results: 
# 1) Do my proteins have more overlaps with promoters or genebodies?



# It is nice and all to find overlaps, but I am interested in how many proteins
# bind a specific promoter. I will use my handy "occurence" parameter in 
# " count peaks per feature" 

```{r}

```
## results: I find the max number of proteins on a promoter to be X


# Now I want to start plotting my results
# First I will see if there is a realtionship between peak number and total DNA covered
```{r}


ggplot ()
```

# Now I want to color my plot by wether the protein is a TF or not.
```{r}


ggplot
```

# I want to make a histogram of the number of peaks for each of my proteins

```{r}



hist
```


# Now I want to facet this by the type of DNA binding domain my protein has.
```{r}

```


# Cool now I am ready to send my result to my collaborator as a
# Knitted document

