---
title: "02_plotting_1"
author: "JR"
date: "8/3/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)

library(ggplot2)
# install.packages("ggplot2")
# install.packages("broom")
library(ggpubr)
library(tidyverse)

source("../../../../util/plotting_functions.R")
source("../../../../util/_setup.R")

# source("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/util/plotting_functions.R")
# source("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/util/_setup.R")

```

One of the best parts of R is the plotting abilities. You are, by analogy designing your own figures with code. You can run many stastical analyses and plot the outputs in the code to make a figure. Here we will focus on GGPLOT(GGPLOT2). The GG says it all:

"Grammar of Graphics" -- it truly is a very pure and flexible way of plotting data for figures!

There are 8 possible layers of information -- each building on the data layer that you are importing. This looks really confusing but we will slowly build up from the basics (mapping, stats, geom)

Plotting layers of R:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

4) scales -- how big small you want things

5) GEOMETRIES -- what type of plots you are making (geom)

6) Facets -- making multiple representations of plots

7) Coordinates -- specifing placement of image aspects

8) THEME -- a set version of colors, background. Mostly so you don't have to add many of the layers above everytime!

Ok let's walk through some simple examples to more difficult figures using these layers!

Layer 1 : DATA (data types matter so good to label object accrodingly)

# loading in num_peaks_df

```{r loading in num_peaks_df}
# before we get started let's import that data we are going to plot from num_peaks_df. 
num_peaks_df <- read_csv('../01_peak_features/results/num_peaks_df.csv')

# this maybe a bit boring so we can try reading in the bigger dataset we will have at end of class
big_num_peaks_df <- read_csv("/scratch/Shares/rinnclass/CLASS_2023/data/data/2022_num_peaks_df.csv")

# num_peaks_df <- read_csv('/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/num_peaks_df.csv')
# num_peaks_df <- num_peaks_df[1:12]
# write_csv(num_peaks_df, "/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/num_peaks_df.csv")

```

# GGPLOT AES Layer and Geom : basic example
Now let's start plotting in GGplot!


```{r ggplot basics}

?ggplot
# ggplot first wants the data parameter, then an 'aesthetics' layer (aes)
# aes() layer needs everything you want to do with the data. The + adds the next layer.


# let's make this simple plot first: number of peaks -vs- total peak length
ggplot(big_num_peaks_df, aes(x = num_peaks, 
                         y = total_peak_length)) +
  geom_point() 
# We have a result! The more peaks the more genome space is covered.
# Let's make a folder in "analysis" 14_ to put our figures in.


# Same as :

big_num_peaks_df %>%
ggplot(aes(x = num_peaks,
           y = total_peak_length)) +
  geom_point()


```

# changing shapes in plot
Now let's change the shape of the points outside the aes() layer and in the geom() layer
```{R changing shapes in plot}
             
ggplot(big_num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
         
           geom_point(shape = 'square',
             color = 'purple')


# Here are some good resources for other shapes and color vignettes:
# http://www.sthda.com/english/wiki/ggplot2-point-shapes

```

# coloring by data in aes() layer
We colored the plot above in the geom layer, but what if we wanted to color by "is_tf"?
Since that depends on the data we need to do that in the aes() layer!
We can even index and run functions in aes()
```{R # coloring by data in aes() layer}
# Let's plot the num_peaks -vs- coverage for Zinc finger proteins (C2H2 ZF)

ggplot(big_num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length,
                 color = dbd == "C2H2 ZF")) +
  geom_point()

```

```{R adding function in aes() layer}
#simple function in the aes()
ggplot(big_num_peaks_df, aes(x = log2(num_peaks/1e3), 
                 y = total_peak_length/1e6,
                 color = dbd == "C2H2 ZF")) +
  geom_point()

# Wow that sure looks differnt :) of note the inflection point
# indicates at a certain point the trend is not as linear as we saw before.

```

# facet in ggplot by TF annotation
We can also plot each type of data in a column by using the 'facet' function.
Let's plot three plots for if DBP is a TF, is not or NA.

```{R facet by TF annotation}

ggplot(big_num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length
                )) +
  facet_wrap(tf ~ .) +
  geom_point() 

# nice we see three plots, one for each category.
# We have a result: All categories have linear relationship
  
```

# Geom layer : histogram
Now let's take a look at other types of plots in the geom layer.
geom_histogram() is very useful in a lot of analyses. Let's start there.

```{r Geom_histogram}

# frequencty of peak numbers across DBPs

ggplot(big_num_peaks_df, aes(x = num_peaks)) +
  geom_histogram()

?geom_histogram()
# Result: peak number is represented by right gaussian.
# let's add more bins

ggplot(big_num_peaks_df, aes(x = num_peaks)) +
  geom_histogram(bins = 30)

# Let's chage the color of the bars in histogram by number of peaks.
# again we need to do this inside aes() since the data needs to read.

ggplot(big_num_peaks_df, aes(x = num_peaks, fill = tf)) +
  geom_histogram(bins = 30)

# this is hard to see the two distributions.
# let's fix this by adding a 'position' parameter in the geom layer.

ggplot(big_num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_histogram(bins = 30, position = "dodge")

# so now we can see the individual data points separated!

```

# Geom layer : density plot
DENISTY Plots are also very useful to see where the density of data arises.
```{R geom_density()}

ggplot(big_num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_density()

# but again its hard to see let's fix that with shading or alpha in
# geom layer

ggplot(big_num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_density(alpha = 0.1)

# let's change color of the line now (inside aes())

ggplot(big_num_peaks_df, aes(x = num_peaks, fill = tf, color = tf)) +
   geom_density(alpha = 0.3)

# so now line and fill are same color -- let's see:

ggplot(big_num_peaks_df, aes(x = num_peaks, fill = tf, color = tf)) +
   geom_density(alpha = 0.0)

# Result TF and other DBPs have similar peak properties

```

# Multiple Geom layers 
The geom layer has a lot of options AND you can layer them!
Let's take out x,y plots above and map them into a 2D density plot
```{R geom_point() & geom_density_2d}

ggplot(big_num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
  geom_point()+
  geom_density_2d() 

# This is helpful as it shows us most of the data is in bottom left

# what would happen if we change the order?

ggplot(big_num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
   geom_density_2d()+
    geom_point()

# it's hard to see but in this case the contour is on top
# of points and points are below -- subtle but important

```

# trend lines
Let's add even more geom_*() layers -- starting with a 'trend line'
```{r geom_smooth, geome_abline}

?geom_abline
# This will make a line with a given slope and intercept

?geom_smooth
# This shows variance and deviations of the trend lines

# let's add them both !

ggplot(big_num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
  geom_point() +
  geom_abline(slope = 1000, intercept = 0) +
  geom_smooth(method = "lm")


?geom_smooth

```

# Bar plots
Let's take a quick look at bar plots 
we provide the x-axis and ggplot counts for Y.
We will also use this opportunity to look at level 3) STAT layer
```{r geom_bar()}

# let's plot the number of each type of DBD in our set of DBPs 

ggplot(big_num_peaks_df, aes(x = dbd)) + 
  geom_bar()


# If the we want to provide the Y-axis for a specific DBP
# we need to apply stat = identitiy in the geom_bar()

ggplot(big_num_peaks_df, aes(
      x = dbp,
      y = num_peaks)) +
        geom_bar(stat = "identity")

# basically this allows you to command the y-axis 
# but is very messy in this case


# let's look at a more realistic example when there is categorical data

ggplot(big_num_peaks_df, aes(x = tf, 
                 y = total_peak_length
                )) +
  geom_point() +
  stat_summary(fun = "mean",
               color = "red",
               shape = 2,
               size = 2)

?stat_summary
# We can use a linear model to find the best trend-line

ggplot(big_num_peaks_df, aes(x = num_peaks, y = total_peak_length)) +
  geom_point() +
  geom_smooth(method = "lm")

# We can get the slope and intercept from making a lm variable:

lm_fit <- lm(big_num_peaks_df$total_peak_length ~ big_num_peaks_df$num_peaks)
summary(lm_fit)
# slope = 1,000 
# ORDER MATTERS: if reversed the slope is 8 e -3!
lm_fit <- lm(big_num_peaks_df$num_peaks ~ big_num_peaks_df$total_peak_length)
summary(lm_fit)


# let's convert the lm_fit data to a data_frame using tidy()
# it's in the library broom:
library(broom)

lm_df <- tidy(lm_fit)
lm_df
# so much easier to read than summary :)

# plotting with the actual lm fit:
#TODO geom_abline doesn't work
ggplot(big_num_peaks_df, aes(x = num_peaks, y = total_peak_length)) +
  geom_point() +
  geom_abline(slope = lm_df$estimate[[2]], intercept = lm_df$estimate[[1]])

# See how this changes if order is changed

```

# Summary of where we are at
Now we have gone through these layers:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

5) Geom

Now let's look at the scales layer

4) scales -- how big small you want things
Scales is important for outputting specific factors as colors that are "scaled"
alos other outputs need to be scaled such as scale values to the "mean" etc.


# scale layer
?scale

```{r scale layer}

# Let's see how we can use the scales layer to change axes.

ggplot(big_num_peaks_df, aes(x = log10(num_peaks), y =log10(total_peak_length), color = tf)) +
  geom_point()+
  scale_x_continuous(breaks = c(1, 2, 3, 4)) +
  scale_color_brewer(palette = 3) 

# scale_x_continious (same for y) is calling the scale layer specifically
# We are giving it where to put values on X-axis      


```

# changes axes limits
What if we want to set the limits of the axes? 
We can use 'xlim' & 'ylim' in scale layer.

```{R xlim and ylim}

ggplot(big_num_peaks_df, aes(x = num_peaks, y = total_peak_length, color = tf)) +
  geom_point() +
  xlim(1e3, 1e5) +
  ylim(5e5, 6e7) +
  scale_color_brewer(palette =  2) 

# coool so we zoomed in!

```

# FACETS
This will allow us to make multiple plots in the same plot frame or facet
One limitation is that it can only make the same plots from different data

Be careful in the facet layer that you don't make too many plots. 
Essentially we make as many figures as there are entries.

FACET_GRID & FACET_WRAP
```{r facet_grid(), facet_wrap()}

# First let's take a look at the top three represented DBDs?
table(big_num_peaks_df$dbd)[order(table(big_num_peaks_df$dbd))]

# Now let's set dbds to these three DBDs.
dbds <- c("C2H2 ZF", "Homeodomain", "bZIP")

# Now the plotting beings and good example of running functions in aes()
ggplot(big_num_peaks_df %>% filter(dbd %in% dbds),
       aes(x = num_peaks, y = total_peak_length )) +
  facet_grid(dbd ~ .) + 
  geom_point()

# here we call facet grid. The parameters are rows and columns separated by ~
# so we said take the DBDs and plot them as rows.
```

We can also facet grid another parameter -- let's say tf

```{R two facets}

ggplot(big_num_peaks_df %>% filter(dbd %in% dbds),
       aes(x = num_peaks, y = total_peak_length )) +
  facet_grid(dbd ~ tf) + 
  geom_point()



# Here we said plot DBD as rows and tf status as columns -- 
# We can easily flip this too:


ggplot(big_num_peaks_df %>% filter(dbd %in% dbds),
       aes(x = num_peaks, y = total_peak_length )) +
  facet_grid(tf ~ dbd) + 
  geom_point()

# Thus the ~ is telling which rows and columns are plotted

```

# facet wrap
We just used facet_grid, but we can also use facet_wrap
```{R facet_wrap()}

ggplot(big_num_peaks_df %>% filter(dbd %in% dbds), 
       aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(dbd ~ tf) + 
  geom_point()

# difference all combos of each factor. Same as before but facet wrap
# doesn't mind if all the values are in one category. 
# Overall the same plots as facet_grid()
```

# DEEPER INTO FACET: we can add some parameters to order the plots
```{R deeper into facets}

ggplot(big_num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., nrow = 3, ncol = 1) + 
  geom_point()

# or 

ggplot(big_num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., nrow = 2) + 
  geom_point()

# Note the axis don't scale here -- let's use 'scales = free' parameter

ggplot(big_num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., scales = "free", ncol = 1) + 
  geom_point()

```

# Coordinate layer
How to represent X and Y coordinates in different coordinate systems.
This is unlikely to really be used in biological data but an important layer!


R has a bunch of built in "coordinate" systems. We have been using Cartesian 
and it is the most common "coordinate" system.
But you can also plot data in "polar" and other coordinate systems.
We can just add it in the "Coordinates layer"

```{r coordinate layer}

# let's make the coordinates polar instead of cartesian
ggplot(big_num_peaks_df, aes(x = dbd)) +
  geom_bar() +
  coord_polar()

# the values eminate from the center and are a "bar_plot" from the center.
# we can easily see as we did above C2H2 ZF most represented in our data.

# We can play around in the coordinate space too. 
# For example to tell what is  the "theta" or circular dimension. 

ggplot(big_num_peaks_df) +
  geom_bar(aes(x = dbd)) +
  coord_polar(theta = 'y')

```

# Theme layer
This is a great way to package all the features you want, colors,
backgrounds, grid lines etc. We tend to use paper white as a defualt theme.

We made a defualt theme that we can source at the begining of the
document call _setup.R. Once loaded a "paper white" theme will be available.

Here is an example of a 'theme' is called in ggPlot

```{r themes}

ggplot(big_num_peaks_df, aes(x = num_peaks)) +
  geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  theme_paperwhite() +
  xlab(expression("Number of peaks")) +
  ylab(expression("Density")) +
  ggtitle("Promoter binding events",
          subtitle = "mRNA and lncRNA genes")

# Note this example build everything from labels to data presentation!

# Let's take a look at _setup.R
# Nice, it basically calls a "function" in ggplot
# You can change this as you please by going through each line (hint: google)
```

# more themes
There are a number of built-in themes in addition to 
the one we've provided. You can see them by typing theme_ 
then autocomplete and check a few out! Here is cleavland :

```{r more themes}

ggplot(big_num_peaks_df, aes(x = num_peaks)) +
  geom_density() +
  theme_cleveland()

# we can load our custom theme paper_white
ggplot(big_num_peaks_df, aes(x = num_peaks)) +
  geom_density() +
  theme_paperwhite()


```
