---
title: "01_to_clone"
author: "JR"
date: "7/20/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(tidyverse)
library(httr)
library(janitor)
library(purrr)
source("/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/util/class_functions.R")
source("/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/util/encode_functions.R")
```

The goal for today is to go to the encode portal and 
choose 2 proteins per student to study. Or more if you want.

Let's go to the portal and pick some out :

1) Assay Title: Click TF seq
2) Target of assay: pick your favorite proteins
3) Biosample: Hepg2
4) Analysis: fastq
5) make sure to click NO on the "hide controls" 
5) download selected files (in dropbown)

# I chose HDACs that are here :
[encode_query](https://www.encodeproject.org/report/?type=Experiment&status=released&perturbed=false&biosample_ontology.classification=cell+line&target.label=HDAC2&target.label=HDAC1&biosample_ontology.term_name=HepG2&biosample_ontology.classification=cell+line&files.file_type=fastq&field=%40id&field=accession&field=assay_term_name&field=assay_title&field=target.%40id&field=target.label&field=target.genes.symbol&field=biosample_summary&field=biosample_ontology.term_name&field=dbxrefs&field=description&field=lab.title&field=award.project&field=status&field=files.%40id&field=related_series&field=replicates.library.biosample.accession&field=replicates.biological_replicate_number&field=replicates.technical_replicate_number&field=replicates.antibody.accession&field=replicates.library.biosample.organism.scientific_name&field=replicates.library.biosample.life_stage&field=replicates.library.biosample.age_display&field=replicates.library.biosample.treatments.treatment_term_name&field=replicates.library.biosample.treatments.treatment_term_id&field=replicates.library.biosample.treatments.amount&field=replicates.library.biosample.treatments.amount_units&field=replicates.library.biosample.treatments.duration&field=replicates.library.biosample.treatments.duration_units&field=replicates.library.biosample.synchronization&field=replicates.library.biosample.post_synchronization_time&field=replicates.library.biosample.post_synchronization_time_units&field=replicates.library.biosample.applied_modifications.modified_site_by_target_id.organism&field=replicates.library.biosample.applied_modifications.introduced_gene.organism&field=replicates.%40id&field=replicates.library.mixed_biosamples&field=replicates.library.biosample.subcellular_fraction_term_name&field=replicates.library.construction_platform.term_name&field=replicates.library.construction_method&field=possible_controls&limit=all)

The above file has all the info to download FASTQ files (the most raw data)
Now let's download the fastq files
```{bash}

# Use a file transfer program to place this in the directory you want to download to
cd /scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/05_enCODE/data/

# now let's download with the command provided by encode:
xargs -L 1 curl -O -J -L < files.txt

# note this is pretty cool. xargs is like bash memory 
# so this is saying open my memory (xargs)
# import a list (files.txt)
# skip one line and use curl to download each of the urls recursively 
```

Now we need the sample information:

1) Click the report button
2) click on download TSV button 
3) rename to samples.txt and put in data dir with file transfer

[report_tsv](https://www.encodeproject.org/report.tsv?type=Experiment&status=released&perturbed=false&biosample_ontology.classification=cell+line&target.label=HDAC2&target.label=HDAC1&biosample_ontology.term_name=HepG2&biosample_ontology.classification=cell+line&files.file_type=fastq&field=%40id&field=accession&field=assay_term_name&field=assay_title&field=target.%40id&field=target.label&field=target.genes.symbol&field=biosample_summary&field=biosample_ontology.term_name&field=dbxrefs&field=description&field=lab.title&field=award.project&field=status&field=files.%40id&field=related_series&field=replicates.library.biosample.accession&field=replicates.biological_replicate_number&field=replicates.technical_replicate_number&field=replicates.antibody.accession&field=replicates.library.biosample.organism.scientific_name&field=replicates.library.biosample.life_stage&field=replicates.library.biosample.age_display&field=replicates.library.biosample.treatments.treatment_term_name&field=replicates.library.biosample.treatments.treatment_term_id&field=replicates.library.biosample.treatments.amount&field=replicates.library.biosample.treatments.amount_units&field=replicates.library.biosample.treatments.duration&field=replicates.library.biosample.treatments.duration_units&field=replicates.library.biosample.synchronization&field=replicates.library.biosample.post_synchronization_time&field=replicates.library.biosample.post_synchronization_time_units&field=replicates.library.biosample.applied_modifications.modified_site_by_target_id.organism&field=replicates.library.biosample.applied_modifications.introduced_gene.organism&field=replicates.%40id&field=replicates.library.mixed_biosamples&field=replicates.library.biosample.subcellular_fraction_term_name&field=replicates.library.construction_platform.term_name&field=replicates.library.construction_method&field=possible_controls&limit=all)

Great now we have:

1) our fastq files downloaded (may take 20-30 min) - from files.txt
2) sample information on each experimental acession.

The experimental accession (ENCSR) is comprised of many experimental files (ENCFF)
The experimental files include the controls (input) and replicates etc.
So there will be many more ENCFFs than ENCSRs


now we are going to use ENCODE API to get the information we need
this was inspired by the lack of md5sum info and others
so we figured out how to find it in the API


# Using ENCODE API
 ---- Introduction to APIs (Application Programming Interface) ----
 
 In order to exchange information between someone's database and your computer you use an API 
Application Programming Interface. Basically a highly specified language for interacting 
and retrieving the data you need.

The ENCODE API provides extensive documentation as to what data you can request 
and how to format that data to make your request. You can browse the possible requests in
an interactive way using their interactive documentation:
https://app.swaggerhub.com/apis-docs/encodeproject/api/basic_search

Now, we will use ENCODE's API to retrieve additional file information from their server.

# establish the API mode of communication
```{r encode API syntax}

# (1) we need the encode URL to communicate with the API
# we see this in the URL we used to download the TSV above

base_url <- "https://www.encodeproject.org/report.tsv?"

# Let's look at an example request for this experiment accession: ENCSR541TIG

request_url <- "https://www.encodeproject.org/report.tsv?type=File&status=released&file_format=fastq&dataset=%2Fexperiments%2FENCSR541TIG%2F&field=accession&field=read_count&field=md5sum&field=controlled_by&field=paired_end&field=paired_with&field=replicate&field=target"

# the field parameter is where we tell it which columns or which pieces of data we want to get.
# this retrieves read_count, md5sum, controlled_by, paired_end, paired_with, replicate, and target
# NOTE API language use :
# file_format=fastq&dataset=%2Fexperiments%2FENCSR541TIG%2F&field=accession&
# dataset=%2F
# Experiments%2F .....


# ok now we have the info we need to connect to API and get what we want !
# we are essentially creating a custom TSV from the one we downloaded 
```



# Now we can create a function to make URLs to get the info we want
# Function 1 : construct query
```{r ENCODE API query function}

# first let's set up the function and it's parameters
contstruct_query <- function(experiment_accession,
                             base_url = "https://www.encodeproject.org/report.tsv?",
                             file_format = "fastq",
                             type = "File",
                             status = "released",
                             fields = c("accession", "read_count", "md5sum",
                                        "controlled_by", "paired_end",
                                        "paired_with", "replicate", "target")) {
  
  # note the function is still open "{ "
  
  # Now we will populate this structure above, note experiment_accession is only
  # parameter we need to populate
  # We are copying the terminology used in REQUEST_URL or communicate with API
  query <- paste(list(paste0("type=", type),
                      paste0("status=", status),
                      paste0("file_format=", file_format),
                      
                      # We are using same language as Encode API that has %2F as separator
                      paste0("dataset=%2Fexperiments%2F", experiment_accession, "%2F"),
                      
                      # map is a way of transforming input and applying a function
                      # in this case we are just using "paste0" as the function
                      # map_chr is to make sure it stays as a character value
                      map_chr(fields, ~paste0("field=", .))) %>%
                   flatten(),
                 collapse = "&")
  url <- paste0(base_url, query)
  return(url)
}
# essentially we just recreated the base URL with addition information 
# in fact using the logic we got Md5 values and they are not accessible on web!

```

# FUNCTION 2: encode_file_info
This function actually makes the request and returns the data only 
(without the response headers) in a data.frame format.

```{R encode_file_info function}
# setting up the function and parameters
# this function will go get the data from the URL we made above
encode_file_info <- function(experiment_accession,
                             base_url = "https://www.encodeproject.org/report.tsv?",
                             file_format = "fastq",
                             type = "File",
                             status = "released",
                             fields = c("accession", "read_count", "md5sum",
                                        "controlled_by", "paired_end",
                                        "paired_with", "replicate", "target")) {
  
  # Now we are creating a url that encode will understand
  path <- "report.tsv?"
  base_url <- modify_url("https://www.encodeproject.org/", path = path)
  # note we are now merging in the function from above "construct query"
  url <- contstruct_query(experiment_accession,
                          base_url = base_url,
                          file_format,
                          type,
                          status,
                          fields)
  # this is now retrieving the data with GET function in httr
  resp <- GET(url)
  if (http_error(resp)) {
    error_message <- content(resp, type = "text/html", encoding = "UTF-8") %>%
      xml_find_all("//p") %>%
      xml_text() %>%
      first()
    stop(
      sprintf(
        "ENCODE API request failed [%s]\n%s",
        status_code(resp),
        error_message
      ),
      call. = FALSE
    )
  }
  
  if (http_type(resp) != "text/tsv") {
    stop("API did not return text/tsv", call. = FALSE)
  }
  body <- read_tsv(content(resp, "text"), skip = 1) %>%
    clean_names()
  return(body)
}
```

Above we just created a function that is in our local environment
So now we can type "encode_file_info(paramters)" and the function will run
But we need an input to the function which is Experimental Acession #

# reading in sample sheet to get experimental acession for YFTF

```{r reading in sample sheet}

# We'll also rename this Accession column to clarify between experiment_accession and file_accession.

samples <- read.table("data/samples.txt",
                      sep = "\t", skip = 1, header = T) %>%
  # just renaming accession to experimental acession
  dplyr::rename(experiment_accession = Accession) 


# We also want to download the control experiments, which means we need to retrieve info about those as well.
controls <- samples %>%
  dplyr::select(Controls) %>%
  rowwise() %>%
  mutate(Controls = gsub("/experiments/|/", "", Controls)) %>%
  separate_rows(Controls, sep = ",") %>%
  dplyr::rename(experiment_accession = Controls)

samples <- bind_rows(samples, controls)


# It's seems mundane but starting here is the best way to make a "reproducible" sample sheet.
# Bottom line: the download to code to analysis is the way to reproducibility (this worked a year later :)

# Nice so now we have a sample sheet with experiment accessions 
# Now we can run encode_file_info function on this file and DONE !

```
# ENCODE sometimes deprocates files and in my case two of the files went missing
# This technically doesn't cause any problems.
# It can be diagnosed by the files not being renamed in 02 
# Also you can check how many lines files.txt -vs samples.txt
# If samples.txt has less lines some files were deprocated 


# running encode function to get our final sample sheet.
```{r MAP to experiment_accession}

# we are using two functions here:
?mutate
# mutate will make a new column and then run function (encode_file_info)
?map
# map will take each experiment accession in "experiment_accession" col
# then pass that into encode_file_info (via the ~) as input .x into encode_file_info
# map will nest all the files for 1 experimental accession number into one "cell"
# map is amazing and we can discuss conceptually -- if you want to play with
# it more there is some bonus code on the bottom of RMD.

# Let's run it : 

samples <- samples %>%
  # note function inside a function being called.
  mutate(file_info = map(experiment_accession, ~ encode_file_info(.x)))
# Thus for each accession in experiment_accession mutate will use the function MAP
# to map each experimental file to the "file_info" column and keep the data there.

# let's see what this looks like:

samples
# 5 obs and 40 vars
# cool we got md5sums we will need in a bit!
```

Now we need to unnest the mapped data

# Unnesting the data from MAP
This is a bit hard to read in this format, so we can unnest the data.frames in the file_info column using the unnest command.
This will duplicate the remaining columns to match the number of lines in the nested data.frame.

```{r unnesting MAP}
?unnest
# This function will extract the mapped info and make new cols and rows
# We also need to tell it which column we want to unnest.
# note samples is currently 5 rows
samples <- samples %>%
  unnest(cols = file_info)

# let's look and see what happened
samples
# Now samples is 12 rows (obs) and 47 vars from unnesting 
# previous 5 rows: each experimental accession had 4 files for 5 exp accession = 20 rows
# Note we went from 48 obs to 47 because the col the samples mapped to were removed with unnest
```

Excellent we now have our sample shee cleaned up and ready to use
to make a design file for our new chip run !

# Write out sample sheet to be loaded in next lesson
# I think a good place for this is where we will run the next pipeline in nextflow:

/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/03_Nextflow/01_my_chipseq

```{r writting out sample sheet}

write_csv(samples, "/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/03_Nextflow/01_my_chipseq/samples.csv")
samples <- read_csv("/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/03_Nextflow/01_my_chipseq/samples.csv")
```

```{r}
# We need to download the control files...
samples <- samples %>%
  mutate(dl_url = paste0("wget https://www.encodeproject.org/files/", accession, "/@@download/", accession, ".fastq.gz"))

ctls <- samples %>%
  filter(is.na(ID))
write_lines(c("#!/bin/bash", ctls$dl_url),
            "data/dl_ctls.sh")

chip <- samples %>%
  filter(!is.na(ID))
write_lines(c("#!/bin/bash", chip$dl_url),
            "data/dl_chip.sh")
```


