---
title: "01_create_consensus_peaks"
author: "JR"
date: "8/10/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
source("../../util/my_class_functions.R")

# filepath to import peaks
basepath <- "/scratch/Shares/rinnclass/CLASS_2023/<your_folder>"
peak_path <- "CLASS_2023/CLASSES/05_R_analyses/analysis/00_consensus_peaks"
broadpeakfilepath <- "/scratch/Shares/rinnclass/CLASS_2023/data/data/peaks"
```

# Loading in all .broadPeak files for all dbps and their replicates

```{r load .broadPeak files}

# using import peaks to import .broadPeak files (~10min)
peak_list <- import_peaks(consensus_file_path = broadpeakfilepath)

```

# creating consensus peaks
Using consensus_from_reduced function that requires peak list
and dbp object. This will be done in 2 steps
(1) created dbp object of unique names for dbp
(2) run consensus_from_reduced


```{r create consensus peaks}

  # Creating unique DBP object for create_consensus_peaks_from_reduced
dbp <- unique(sapply(names(peak_list), function(x) {
   unlist(strsplit(x, "_"))[1]
}))

# now run our function consensus_from_reduced
consensus_list <- lapply(dbp, consensus_from_reduced, peak_list)

# adding names to the GRange list
names(consensus_list) <- dbp

save(consensus_list, file = "results/consensus_list.RData")

load("results/consensus_list.RData", verbose = T)
```

# exploring the number of peaks in the consensus_list
```{r exploring number of peaks in consenus peaks}

# creating list of num_peaks per dbp
num_peaks <- sapply(consensus_list, length)

# plotting
hist(num_peaks, breaks = 1000)
hist(num_peaks, breaks = 1000, xlim = c(0,3000))

```
Result: seems like 1000 peaks should be the min moving forward.

# filtering consensus_list to dbps with > 1000 peaks
```{r filtered_consenus_list}

# filtering to 1000 peaks
filtered_consensus_list <- consensus_list[sapply(consensus_list, length) > 1000]

# saving 
save(filtered_consensus_list, file = "results/filtered_consensus_list.RData")

# keeping track of DBPs lost
lost_dbps <- names(consensus_list[sapply(consensus_list, length) < 1000]) %>% as.data.frame()

# same as
# lost_dbp <- names(consensus_peaks)[!(names(consensus_peaks) %in% names(filtered_consensus_peaks))]

# saving 
write.table(lost_dbps, "results/lost_dbps.csv")

```

# exporting filtered_consensus_peaks
```{r exporting filtered consensus peaks}

for(i in 1:length(filtered_consensus_list)) {
  rtracklayer::export(filtered_consensus_list[[i]], 
                      paste0("results/filtered_consensus_peaks/", 
                             names(filtered_consensus_list)[i], 
                             "_filtered_consensus_peaks.bed"))
}

#TODO rename so can load into UCSC.


```

# loading in genome features
```{r creating genome feature objects}

gencode_gr <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2023/data/data/genomes/gencode.v32.annotation.gtf")

# gencode genes
gencode_genes <- gencode_gr[gencode_gr$type == "gene"] 

# mrna_genes
mrna_genes <- gencode_genes[gencode_genes$gene_type %in% "protein_coding"]

# lncrna_genes
lncrna_genes <- gencode_genes[gencode_genes$gene_type %in% "lncRNA"] 

# mrna_lncrna_genes
mrna_lncrna_genes <- gencode_genes[gencode_genes$gene_type %in% c("protein_coding","lncRNA")]

# lncrna_mrna_promoters
lncrna_mrna_promoters <- promoters(mrna_lncrna_genes, upstream = 1000, downstream = 1000)

# lncrna_gene_ids
lncrna_gene_ids <- mrna_lncrna_genes$gene_id[mrna_lncrna_genes$gene_type == "lncRNA"]

# mrna_gene_ids
mrna_gene_ids <-mrna_lncrna_genes$gene_id[mrna_lncrna_genes$gene_type == "protein_coding"]

```


# making data frame of filtered_consensus_peak info
```{r creating num_peaks_df to track peak properties}

num_peaks_df <- data.frame("dbp" = names(filtered_consensus_list),
                           "num_peaks" = sapply(filtered_consensus_list, length))

# total genome covered by peaks
num_peaks_df$total_peak_length <- sapply(filtered_consensus_list, function(x) sum(width(x)))

# creating number of promoter overlaps entry
promoter_peak_counts <- count_peaks_per_feature(lncrna_mrna_promoters, filtered_consensus_list, type = "counts")

# creating promoter peak_occurence for clustering - Metaplots later.
promoter_peak_matrix <- count_peaks_per_feature(lncrna_mrna_promoters, filtered_consensus_list, type = "occurrence")

# saving
write.table(promoter_peak_matrix, "results/promoter_peak_occurrence_matrix.tsv")

# read back in
promoter_peak_occurrence_matrix <- read.table("results/promoter_peak_occurrence_matrix.tsv")

# summing rows to get total number of promoter overlaps
num_peaks_df$peaks_overlapping_promoters <- rowSums(promoter_peak_counts)

# lncrna promoter overlaps 
num_peaks_df$peaks_overlapping_lncrna_promoters <- rowSums(promoter_peak_counts[,lncrna_gene_ids])

# mrna promoter overlaps
num_peaks_df$peaks_overlapping_mrna_promoters <- rowSums(promoter_peak_counts[,mrna_gene_ids])

# Finding overlaps with gene_bodies (will take a few minutes again)
# Note this takes several minutes
genebody_peak_counts <- count_peaks_per_feature(mrna_lncrna_genes, 
                                                filtered_consensus_list, 
                                                type = "counts")

# All gene bodies overlaps
num_peaks_df$peaks_overlapping_genebody <- rowSums(genebody_peak_counts)

# lncRNA gene bodies 
num_peaks_df$peaks_overlapping_lncrna_genebody <- rowSums(genebody_peak_counts[,lncrna_gene_ids])

# mRNA gene bodies
num_peaks_df$peaks_overlapping_mrna_genebody <- rowSums(genebody_peak_counts[,mrna_gene_ids])


```

# adding TF type annotations for DBPs
```{r adding TF features to num_peaks_df}

# download TF annotations to results
url <- "https://www.cell.com/cms/10.1016/j.cell.2018.01.029/attachment/ede37821-fd6f-41b7-9a0e-9d5410855ae6/mmc2.xlsx"
destination_for_url <- "results/TF_annotations.xlsx"

# to download we can use download.file
download.file(url, destination_for_url)

# reading in TF annotations 
human_tfs <- readxl::read_excel("results/TF_annotations.xlsx",
                                sheet = 2, skip = 1)

# let's rename the 4th column to indicate if it is a TF.
names(human_tfs)[4] <- "is_tf"

# now let's intersect gene names that are in our ChIP data and has TF identity.
length(which(tolower(num_peaks_df$dbp) %in% tolower(human_tfs$dbp)))
# 407 of the 430 have matching gene_names - not bad

human_tfs <- human_tfs[tolower(human_tfs$Name) %in% tolower(num_peaks_df$dbp), 1:4]
# adding new column names
names(human_tfs) <- c("ensembl_id",
                      "dbp",
                      "dbd",
                      "tf")

# merging into num_peaks_df
num_peaks_df <- merge(num_peaks_df, human_tfs, all.x = T)


```


# creating promoter peak occurence matrix
This will make a matrix where promoters are cols (30K)
Each will have 1 if overlapped by a given dbp : 0 if no overlaps

```{r promoter peak occurence matrix}

# running count_peaks_per_feature
promoter_peak_occurence <- count_peaks_per_feature(lncrna_mrna_promoters, filtered_consensus_list, 
                                               type = "occurrence")

# Let's double check that all lncrna & mrna genes are accounted for:
stopifnot(all(colnames(promoter_peak_occurence) == lncrna_mrna_promoters$gene_id))

# saving
write.table(promoter_peak_occurence, "results/lncrna_mrna_promoter_peak_occurence_matrix.tsv")

# Now let's use the 'data.frame()' fucntion. Set up a bunch of colnames and populate them.
peak_occurence_df <- data.frame("gene_id" = colnames(promoter_peak_occurence),
                                "gene_name" = lncrna_mrna_promoters$gene_name,
                                "gene_type" = lncrna_mrna_promoters$gene_type,
                                "chr" = lncrna_mrna_promoters@seqnames,   
                                "1kb_up_tss_start" = lncrna_mrna_promoters@ranges@start,
                                "strand" = lncrna_mrna_promoters@strand,
                                "number_of_dbp" = colSums(promoter_peak_occurence))

# saving
write_csv(peak_occurence_df, "results/peak_occurence_dataframe.csv")


```

# now make a promoter data_frame that tells which dbps are bound
```{r dbp centric promoter occurence}

# dbps on promoters object
DBPs_on_promoter <- lncrna_mrna_promoters %>%
                    as.data.frame() %>%
  dplyr::select(gene_id, gene_name)

# creating promoter dbps by pivot longer of promoter_peak_occurence_matrix
promoter_dbps <- promoter_peak_occurence %>%
  as.data.frame() %>%
  rownames_to_column("dbp") %>%
pivot_longer(2:ncol(.), names_to = "gene_id", values_to = "occurrence") %>%
  filter(occurrence == 1) %>%
  dplyr::select(-occurrence) %>%
  left_join(DBPs_on_promoter)

# checking Firre promoter
firre_promoter <- promoter_dbps %>%
  filter(gene_name == "FIRRE")

# XIST promoter (should be off since male cells)
XIST_promoter <- promoter_dbps %>%
  filter(gene_name == "XIST")

# GAPDH
GAPDH_promoter <- promoter_dbps %>%
  filter(gene_name == "GAPDH")


#TODO pivot longer to make a handy data frame
promoter_dbps_df <- promoter_dbps %>% as.data.frame()
write.csv(promoter_dbps, "results/promoter_dbps.csv")

```

# saving key env objects
```{r saving key objects in environment to load next session}

save(filtered_consensus_list, gencode_genes, lncrna_gene_ids, mrna_gene_ids, num_peaks_df, peak_occurence_df, promoter_peak_occurrence_matrix, lncrna_mrna_promoters, mrna_lncrna_genes, file = "results/peak_features.RData")
# awesome - now we never have to load in consensus peaks again :)

```





