---
title: "02_R_basics"
author: "JR"
date: "7/20/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
```

Today we are going to do some basic R commands and look at our Pol2 ChIP data.
This lesson can be applied to most data frames (similar to excel sheets). 

We will use read_tsv to load in a peak file that is in tab separated format (tsv)
```{bash}

# The Peak files are located in our 03 nextflow folder -- let's take a look:
cd /scratch/Shares/rinnclass/CLASS_2023/<your_folder>/CLASS_2023/CLASSES/03_Nextflow/00_Pol2_NF_CORE_RUN/results/bwa/mergedLibrary/macs/broadPeak

# We see a lot of different types of peak files and we want .braodPeak Let's load one in 

```

Since that is a lot of path typing each time let's shorten it my using file.path R command
```{r file.path}

# change <your_folder> to your folder name to set basepath variable

# we will first need to set an object name to load into the environment (what we are calling this file.path)
# we assign it the values in the file using the assignment opperator " <- " 
basepath = "/scratch/Shares/rinnclass/CLASS_2023/<your_folder>"
#basepath <- "/scratch/Shares/rinnclass/CLASS_2023/JR"

# we can then add the rest of the path as such:
Pol2_peak1 <- read_tsv(file.path(basepath, " rest of file path"))



# now we can read in the peak file using file.path() function , rest of path
# we will use the "read_tsv" baseR function and assign it ( <- ) to Pol2_peak1

Pol2_peak1 <- read_tsv(file.path(basepath, "CLASS_2023/CLASSES/03_Nextflow/00_Pol2_NF_CORE_RUN/results/bwa/mergedLibrary/macs/broadPeak/POLR2A_R1_peaks.broadPeak"))
# cool, let's take a look at what we got by clicking the object in the env window.
```

There are two weird things to note right away:

1) The colnames are the values from the table !
Solution: figure out more about read_tsv

```{r make empty cols}

# let's take a look at the documentation for read_tsv and see if there is a parameter we can use.
?read_tsv

# cool so there is a col_names = F we can use -- let's try it.
Pol2_peak1 = read_tsv("/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/03_Nextflow/00_Pol2_NF_CORE_RUN/results/bwa/mergedLibrary/macs/broadPeak/POLR2A_R1_peaks.broadPeak", col_names = F)

Pol2_peak2 <- read.table("/scratch/Shares/rinnclass/CLASS_2023/JR/CLASS_2023/CLASSES/03_Nextflow/00_Pol2_NF_CORE_RUN/results/bwa/mergedLibrary/macs/broadPeak/POLR2A_R1_peaks.broadPeak", header = F)

write_tsv(Pol2_peak1, "test_pol2.csv")
# Nice so now we have X1-X9 as col names
```

2) There are weird chromosome names that are mislocated DNA "contigs" 
Solution: filter them out.

Let's first rename the cols. Well we need to know what they are :) 
So let's look up the labels of a .broad peak:

# Chromosome, start, stop, name, score, strand, signalValue, pValue, qValue
Signal value = overall enrichment of peak over input
score is how dark the peak will show up in browser (not very useful :)

Ok now we are ready to rename the cols.
There are so many ways to do this we will just explore a couple
```{r renaming columns}

# first let's just change the name of one col (we have X1-X9)
names(Pol2_peak1)[1] <- 'chromosome'

# ok changed the right col name, but now the other cols have no names !!
# not to worry we can index them and rename

# here we used names and then outside that indexed col 2 with [] 
names(Pol2_peak1)[2] <- 'start'


# Load your peak file 
read_tsv("/scratch/Shares/rinnclass/CLASS_2023/...........")

# This will get really repetitive but works. Here is a better solution:
names(Pol2_peak1) <- c('chromosome', 'start', 'end', 'name', 'score', 'strand', 
                       'signalValue', 'pValue', 'qValue')

# here C makes a concatenated list of values -- trick is to make sure in right order!!
```

Cool so now we have a nice organized column names.

# Now let's filter out the weird chromosomes. 
# There are many ways to do this we will test a few

```{r filtering to cannonical chromosomes}

# We are going to first use some "indexing" we will do this A LOT !!
# we keep same object name, we use that object to index using []

Pol2_peak1 <- Pol2_peak1[Pol2_peak1$chromosome %in% c(paste0("chr", 1:22), "chrM", "chrX", "chrY"),]

# Note we have an empty , at the end -- this is because R reads as rows, cols 
# so the empty , is just saying keep all the cols and change the rows 
# also note the use of tidyverse %in% which is like filter -- item has to be "IN" this list.


# here we can do the same thing with less indexing using tidyverse "filter" function

Pol2_peak1 <- filter(Pol2_peak1, chromosome %in% c(paste0("chr", 1:22), "chrM", "chrX", "chrY"))

# tidyverse is awesome because it can now what is inside a data frame with out indexing.
# note we just say "chromosome" to call that col. Above more bulky Pol2_peak1$chromosome


# Let's get even more tidy:

Pol2_peak1 <- Pol2_peak1 %>%
  filter(chromosome %in% c(paste0("chr", 1:22), "chrM", "chrX", "chrY")) 

# Here we use the tidyverse pipe "%>%" to send the information in dataframe (DF)
# to filter and then add the parameters in filter

```

Nice, so now we have a data frame we can start to analyze

# Let's check out some basic features:

```{r basic summary of data frame}

# how many peaks are there -- well it's kinda obvious from env
# but we can use baseR to determine the same value:

length(Pol2_peak1)

# this tells us there are 9 cols

length(Pol2_peak1$chromosome)

# this tells us the chromosome col has 33,493 peaks.


# let's try using summary and table baseR functions;

summary(Pol2_peak1)

# Woah lots of information. Each column is summarized in quartiles etc
# we can see the Pvalues range from 10^-3.401 (0.0003971915) to 10^-154.989 (1.025652e-155)
# qvlaue min is 10^-1.770 (0.01698244) > so MACS uses a 0.01 qValue cut of for sig peaks


# Now let's try table to see how many times a qValue is the same

table(Pol2_peak1$qValue)

# yikes that is hard to understand. But we can table(table :)

table(table(Pol2_peak1$qValue))

# so we can see 1 qvalue shows up 229 times but most have a unique qValue

# let's see if fold change is as variable 
table(table(Pol2_peak1$signalValue))

```

Now let's make a new column that has the width of the peak using "mutate" function

```{r mutate}

# let's see what mutate needs
?mutate

# we will use tidyverse to pipe the DF to mutate so it knows col names

Pol2_peak1 <- Pol2_peak1 %>%
  # mutate wants name of new col and then function to create that col
  # we will simply subtract the end and start of the peak.
  
  mutate(peak_width = end - start)
# That is very handy and a function we will use a lot !!

```


Let's get some info make some basic plots to get to know our data better:

```{r getting to know the peak file}

# let's use summary again:

summary(Pol2_peak1$peak_width)
# cool we see smallest peak is 114 bp and longest 12.4 Kb !

# let's make a histogram plot of peak size distributions using the base R hist func:
?hist

# plotting peak_width col as histogram (we will go into plotting more later in class)
# for now just the basics:

hist(Pol2_peak1$peak_width)

# looks like most are small, let's add more breaks

hist(Pol2_peak1$peak_width, breaks = 50)

# now let's set the x axis limits to ~3Kb

hist(Pol2_peak1$peak_width, breaks = 50, xlim = c(0,3000))

# nice so we observe a left centered normal dist of sizes.

# let's try a density plot -here we need to calculate density then plot:

density <- density(Pol2_peak1$peak_width)

# now we can use plot to see the values of density
plot(density, xlim = c(0,3000))

# here we can see some sort of "shoulder" around 500bp

```

Now let's see how pval correlates with peak size using ggplot
```{r}

# ggplot wasnts to the axis values (peak_width and qValue)
# then it needs to know what kind of plot (we will do much more on this later)

# first pipe in the dataframe to GGplot
Pol2_peak1 %>%
  # set the aes(x= , y= ) 
ggplot(aes(x = peak_width, y = qValue)) +
  # + tells ggplot to add another aspect to plot (points)
  geom_point()

# Interesting, we see that the smaller the peak the lower the qValue tends to be 

```

##################################################
Exercise: compare properties of other replicates
##################################################

# Now load in the other three replicates of Pol2 ChIP
# Address the following

1) What are the average widths of peaks across replicates.
2) What are the number of peaks across replicates
3) What is the length of the smallest and biggest peak in each replicate


##################################################
Exercise: compare properties of other replicates
##################################################

# Do the same as above for the replicates of your groups ChIPseq




