---
title: "04_Iranges-Granges"
author: "JR"
date: "7/21/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(GenomicRanges)
library(tidyverse)
library(IRanges)
# library(Gviz)
```

Iranges is great if there is only one chromosome in the genome :) 
However, there are multiple chromosomes in the genome and there are two strands
of DNA that we need to keep track of.


The GenomicRanges package was developed to include the addition information of chr1 
and other aspects associated with a given interval in IRanges. 
Essentially GenomicRanges is a genome specific version of IRanges. 

** seqnames is the index for the chromosome number


# Genomic Ranges (Granges)
```{R GRanges}

# Let's create a GRange using IRange logic. What we need is:
# Chromosome, Strand, Ranges

gr <- GRanges(seqnames = c("chrX"), strand = c("+", "-", "*"), ranges = IRanges(start = c(1,200, 1000), end = c(30, 299, 1200)))

# let's see 
gr

# the first thing to note is we see a unique identifier for the X chromosome as
# 'seqnames'(similar to names above)
# 'ranges' (same as Iranges)
# 'strand' 

# we can see all the information with seqinfo
seqinfo(gr)

# we can see some additional place holders we have not changed such as the 
# genome version or if the chromosome is circular. Seqlenght is the lenght of 
# the chromosome. let's add this stuff !

seqlengths(gr) <- c("chrX" = 100000)
seqinfo(gr)

# accessing seq lengths:
seqlengths(gr)

# we see that we only have one 'level' here for ChrX, let's add another chromosome
seqlevels(gr) <- c("chrX", "chrY")
seqlevels(gr)
# now we can see that we have a level that contains both the X nad Y chromosome.


# let's add genome:
genome(gr) <- c("hg38")
gr
# now we have a genome, chromosome and strand associated with each feature!

```


Now let's read in two peak files and find their overlaps!
I have some set up to test in the data folder.

# finding overlapping peaks in our actual data sets !!
```{R find overlaps}

# setting file path to peak files
basepath <- "/scratch/Shares/rinnclass/CLASS_2023/JR"
path <- "CLASS_2023/CLASSES/03_Nextflow/00_Pol2_NF_CORE_RUN/results/bwa/mergedLibrary/macs/broadPeak"

# first we read the peak files in as gRanges object with rtracklayer function.
peaks1 <- rtracklayer::import(file.path(basepath,path, "POLR2A_R1_peaks.broadPeak"))
peaks2 <- rtracklayer::import(file.path(basepath,path, "POLR2A_R2_peaks.broadPeak"))

# let's take a look:
peaks1

# for reference to last class looking at this file
peaks_1_tsv <- read_tsv(file.path(basepath,path, "POLR2A_R1_peaks.broadPeak"))

# now we can use the findOverlaps function to see what overlaps!
?findOverlaps
?import

ovf <- findOverlaps(peaks1, peaks2)
length(peaks1)
length(peaks2)
length(ovf)

```
# NOTE: the first file is query and second is the subject. 
# Thus returns how many peak2 overlaps there were in peaks1 as written above.

Let's take a look at the overlaps results.
```{r digging into findOverlap results}
summary(ovf)
# 28445 overlaps -- same result as length
summary(peaks1)
# we had 33582 peaks in replicate 1
summary(peaks2)
# 43435 peaks in replicate 2
# This means 28,445 peaks overlapped but we need to consider 
# the subject and query.
```


# Thought exercise with real data 
# @to and @from 
Imagine the scenario where you had more overlaps than one of 
the peak files had peaks -- mind twister but happens

```{r crazy peaks}
# loading in a situation to get to know query subject (@to and @ from)
crazy_peaks1 <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2023/data/data/POLR2A_consensus_peaks_filter.bed")

crazy_peaks2 <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2023/data/data/POLR2B_consensus_peaks_filter.bed")
```

# If not in class can also download here and import with same code above.
```{bash download crazy peaks}
wget https://www.dropbox.com/s/jenet8pg3t9npls/POLR2A_consensus_peaks_filter.bed?dl=0
wget https://www.dropbox.com/s/vwhpecqxskbfuh4/POLR2B_consensus_peaks_filter.bed?dl=0
```

# Let's find overlaps between these two data sets!
```{r bash download crazy peaks}
# find overlaps
crazy_ovf <- findOverlaps(crazy_peaks1, crazy_peaks2)

# seeing how many overlaps
summary(crazy_peaks1)
length(crazy_peaks2)
summary(crazy_ovf)
# What?? 11042 overlaps but there were only 10,255 peaks in crazy_peaks1
```

# How did we get more than 10,255 peak overlaps? Here are some possibilities:
(1) It could be that all the peaks overlapped and maybe some did more than once
(2) It could be that very few peaks in sample 1 but overlapped numerous times.

These are very different observations!
Let's learn some more Granges while solving this (a common activity)

# Test 1 : width of peaks
```{r digging into findOverlap results}
# let's see with iranges "width function"
summary(width(crazy_peaks1))
summary(width(crazy_peaks2))

# there peaks in peak1 are generally larger than peak2
# thus a peak in peak1 could overlap multiple peaks in peak2 (vice versa too).

```
Result: peaks in crazy_peaks1 are about 2x bigger (consistent with multiple overlaps)


# Test 2: Where did the overlaps come from? Sample (subject) 1 or 2 (query?
We can get more information from then results of findOverlaps.
@from : peaks from query file indexed into subject file
@to: peaks from subject file indexed into query file.

@from ---------       -----  -----

@to  * * *       *                    *

@from = 1 1 1
@to = 1 2 3 

```{r digging into findOverlap results}

# let's understand this more deeply there are some subtilties we should cover
# let's table the @from index
table(crazy_ovf@from) 
table(table(crazy_ovf@from))

# SOLUTION !! This solves it the sum of these is 11K overlaps!
# There are 6,388 unique peak overlaps -- maybe the result you
# actually want to report

# same for @to
table(table(crazy_ovf@to))
# huh most only have one overlap probably because:
# the multiple overlaps in peak2 are assigned to 1 peak in peak1
# 2 peaks in peak2 had overlap with 23 peaks in peak1

```
Result: crazy_peaks1 has much wider peaks and in turn more overlaps than peaks.
Thus Crazy_peaks1 had 6,388 unique peaks and nearly half the overlaps came from
peaks with multiple overlaps in crazy_peaks1. We can also conclude that the smaller 
peaks in crazy_peaks2 were close by to each other -- causing the extra overalps.


# Solution 2: count overlaps
We can look at how overlaps are reported using: countOverlaps

```{r countOverlaps for GRanges}

# let's try countOverlaps
crazy_ovf2 <-countOverlaps(crazy_peaks1, crazy_peaks2)
# NOTE - INTEGER VALUE IN ENVIRONMENT 
# There as many enteries as peaks in crazy_peaks1 (subject)
# each entry has number of times that peak had an overlaps
sum(crazy_ovf2)
# cool this get's us to our 11,042 overlaps!


# let's look with table
table(crazy_ovf2)
# if we sum the overlaps we get the 11,000 !

# crazy_ovf2
#   0    1    2    3    4    5    6    7    8    9   10   11   12   14 
# 2242 6388  918  382  155   81   37   27    8    8    3    4    1    1 

# 6388 + 2*(918) + 3(382) + 4*(155) ...... >> 11,042

```


*********************
EXCERCISE
*********************

Find overlaps for replicates of your group's proteins. 
How do they vary? Peak # differences?

