---
title: "06_functions_1"
author: "JR"
date: "7/31/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(GenomicRanges)
library(tidyverse)
library(IRanges)
# library(Gviz)
# source("../../util/my_class_functions.R")
```
Accompanying Youtube video:
https://youtu.be/56ZI6htKFns


Today we will go over a "function" in R. This is a really nice feature of 
defining a function for R that you may run frequently. 

We have used many of R's built in base functions that require parameters too.
For example,table and summary we used before requires a parameter of an object to summarize etc. 
But the cool part is you can make your own functions just like these base R funcitons. 

Now, let's see what it looks like in R. A common function in life is math 
-- it takes inputs and maps it to outputs.

```{r first function}
# This function has two parameters, and we have to let R know (x & y)
# function is inside { }, then we need a 'return' to get the answer
fun <- function(x, y) {
  ans <- x * y
  return(ans)
}
#let's try it 
fun(2,-4)
# Note that the object ans doesn't exist in our environment. 
# ans is a local variable that is erased after used. global variables are in our environment.
# It's good to remember the "scope" of vars in a function don't really exist outside the function.
# Finally note a function can only return one thing!

```

When creating a function, you can also add documentation about what that function 
does and the data types and expectation for the the parameters it takes.

```{r documenting functions}
#' A function to multiply two numbers
#'
#' @description 
#' This function will multiply the input values of X and Y
#' 
#' @param x one number you'd like to multiply
#' @param  y the other number you'd like to multiply
fun <- function(x, y) {
  ans <- x * y
  return(ans)
}


```

# Let's make this a function we can source from our util directory
(1) go to the util (CLASS_2023/util) and set as working dir
(2) Rstudio: File > newfile > Rscript : paste in function above.
(3) save as "my_class_functions.R" {MAKE SURE you see in util}
(4) go back to /04_R_basics and set as working directory.
(5) Clear your environment -- let's start fresh

#  source("../../util/my_class_functions.R")

# Now let's clean our environment - source this function & - run 
```{r clean env - load packages}
library(GenomicRanges)
library(tidyverse)
library(IRanges)
source("../../util/my_class_functions.R")
# cool we see it loaded in environment and ready to use!
# now type fun(7,6) in console
# We will store all our custom function in this .R script

```


Now let's understand "scope" better by creating a more complex function. 

Local variables are those variables that exist only within a certain part of a program 
like a function and are released when the function call ends.

local variables are mapped to outputs through a function, but only exist within a function.

# Let's make a function to import our Chipseq peak files and extract information from the file name.
This is a very common first step and now we can automate it with some functions!

# CREATE "import_peaks" function
```{r base file path to peaks}

# First let's set a file path to where all our peak files are.
# This is very helpful in the long run to not have huge filepaths everywhere :)
basepath <- "/scratch/Shares/rinnclass/CLASS_2023/JR"
peak_path <- "CLASS_2023/CLASSES/03_Nextflow/00_my_chipseq/results/bwa/mergedLibrary/macs/broadPeak"
broadpeakfilepath <- file.path(basepath, peak_path)


# Now let's make a list of filepaths to each peak file -- "list.files"
# we add a pattern parameter to just load .broadPeaks as there are many other types.

fl <- list.files(broadpeakfilepath, 
                 full.names=TRUE, pattern = ".broadPeak")

# let's see 
fl

# This gives us an index 1:n where each peak file path is a character vector.
# let's try this to grab just one protein (can be the one from your group)
fl_ep300 <- fl[grep("EP300", fl)]

# cool now we can subset this to any protein of our choosing

```

# extracting text from file name (e.g., DBP name)

Let's look at some indexing and ways we can extract file info.
```{R extraction and regular expression} 

# We can index into our file list using [ ]
# let's see what file path 7 is:
fl[7]
# now let's just get the end of the file name that contains the DBP name

# regex is seperating by "/" in file path and taking end $
# regex is intimidating but very helpful. Here is a website that is helpful:
# https://regexr.com/

y <-  str_extract(fl[7], "([^\\/]+$)")

# let's see the difference:
fl[7]
y

# now we can further reduce this to just the DBP name
# string split will split on _ and DBP name is in position 1 after split
unlist(strsplit(y, "_"))[[1]]

# what happens if we try and run together:

y <-  str_extract(fl[28], "([^\\/]+$)") %>%
  unlist(strsplit(y, "_"))[[1]]

# Error !! Why ?
# This is because y is not a local variable. y is global and not set.
# to fix this we need to set y as a local variable in a function
  
```

# setting local variable in function

```{R sapply and regular expression} 

# first we set an object dbp_name 
# next we create a genetic function (here it is generic function x)

dbp_name <- sapply(fl, function(y){
 y <-  str_extract(y, "([^\\/]+$)")
 # NOTE alternative
 # gsub("_peaks.broadPeak", "", y)
  unlist(strsplit(y, "_"))[[1]]
  
})

# let's see what we got
dbp_name
# note in the environment just the DBP name 
# ?? What would the input be to make this a function ??
  
```

# write function to import and rename a list of peak files
let's make this into a real function that is re-usable "import_peaks" and name file by DBP

```{R function to rename files by DBP}

# establishing the function using "function()"
# establishing the parameter consensus_file_path is needed
# extracting TF name as a variable along the way (check env)
import_peaks <- function(consensus_file_path = peak_path) {
  
  # Setting some variables needed in main part of function (same as above -- peak_files & tf_name)
  peak_files <- list.files(consensus_file_path, full.names = T, pattern = ".broadPeak")
  
  # Make an object with each TF name for indexing and merging later
  dbp_name <- sapply(peak_files, function(x){
    y <-  str_extract(x, "([^\\/]+$)")
    unlist(strsplit(y, "_"))[[1]]
  })
  
}

# Nothing happened ?? Becuase we have simply made import peaks function available to us.
# we did not actually run it and it's now in our local environment as a function to be called.
# let's test it:
test <- import_peaks(broadpeakfilepath)

# nice that is working!
```


# final import peaks function
We will use our first for loop !
This will read in each peak file and make a 
list of Granges for each peak file.

```{r full import peaks function}

# setting up the function: all what we just wrote above:
# Get's consensus peak file path, select .broadPeak and then get dbp name from file name
import_peaks <- function(consensus_file_path = broadpeakfilepath) {
  peak_files <- list.files(consensus_file_path, full.names = T, pattern = ".broadPeak")
  dbp_name <- sapply(peak_files, function(x){
    y <-  str_extract(x, "([^\\/]+$)")
    # NOTE ALTERNATIVE Gsub
    # gsub("_peaks.broadPeak", "", y)
    paste(unlist(strsplit(y, "_"))[c(1,2)], collapse = "_") 
    
    })
  
# NOW FOR LOOP TO IMPORT FILES
# setting an empty list "peak_list" to be filled by for loop
  peak_list <- c()
  
  # the for loop !
  for(i in 1:length(peak_files)) {
    # Import peaks
    peaks <- rtracklayer::import(peak_files[i])
    # Append this GRanges object to the of the list.
    peak_list <- c(peak_list, peaks)
    # Name the list elements by their TF name.
    names(peak_list)[length(peak_list)] <- dbp_name[i]
  }
  return(peak_list)
}

# what happened? Nothing ? Why ?

```

# excellent we now have a function that just needs a file path.
# In turn it will load all your peak files into GRanges (imagine doing manually :)
Let's run it 

```{r}

# look how nice and simply this is now!
my_peaks <- import_peaks(consensus_file_path = broadpeakfilepath)

# voila all peak file are loaded 
```


# Now let's make this a function in our util, clean env and start over !

```{r import peaks function }

library(GenomicRanges)
library(tidyverse)
library(IRanges)
source("../../util/your-functions.R")

basepath <- "/scratch/Shares/rinnclass/CLASS_2023/<your_folder>"
peak_path <- "CLASS_2023/CLASSES/03_Nextflow/00_my_chipseq/results/bwa/mergedLibrary/macs/broadPeak/"
broadpeakfilepath <- file.path(basepath, peak_path)

my_peaks <- import_peaks(consensus_file_path = broadpeakfilepath)

# let's see how many peaks are in each file with 1 line!

num_peaks <- sapply(my_peaks, length) %>% as.data.frame()

```
Nice about 4 lines of code to import all desired peak files 
We will use this for thousands of peaks near the end of class
# Conratulations on your new functions :)

*********************
EXCERCISE
*********************

Create a function in your class_functions.R file


