---
title: "01_peak_features"
author: "JR"
date: "8/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(GenomicRanges)
source("/scratch/Shares/rinnclass/CLASS_2022/<your_folder>/CLASS_2022/util/intersect_functions.R")
 # source("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/util/intersect_functions.R")
```

# Exploring the genome features bound by DBPs
Today we are going to explore where in the genome the 483(460) DBPs bind.

From now on we will import all the data we made in "11_creating_consensus_peaks"
Let's start importing all the files we need:


```{R start point 3/11 class}

library(tidyverse)
library(GenomicRanges)
source("/scratch/Shares/rinnclass/CLASS_2022/<your_folder>/CLASS_2022/util/intersect_functions.R")

# first create a file list of filtered consensus peaks
 filtered_consensus_peaks_files <- list.files("/scratch/Shares/rinnclass/CLASS_2022/data/filtered_consensus_peaks", 
                                             pattern = "*.bed",
                                             full.names = TRUE)


# # lappy to import each file in file list 
filtered_consensus_peaks <- lapply(filtered_consensus_peaks_files, rtracklayer::import)

# adding names to rows
names(filtered_consensus_peaks) <- gsub("/scratch/Shares/rinnclass/CLASS_2022/data/filtered_consensus_peaks/|_filtered_consensus_peaks.bed","", filtered_consensus_peaks_files)


# Now gencode annotations
gencode_gr <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2022/data/genomes/gencode.v32.annotation.gtf")

# Finally nump_eaks_df
num_peaks_df <- read_csv("FILE_PATH_12/num_peaks_df.csv")

# num_peaks_df <- read_csv("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/num_peaks_df.csv")

```

# importing genome feature files
Now we will import the GRanges of genome annotations:
```{R importing genome featgures}

# lncRNA and mRNA promoters
lncrna_mrna_promoters <- rtracklayer::import("file_path/lncRNA_mrna_promoters.gtf")

lncrna_mrna_promoters <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/11_consensus_peaks/gene_annotation_files/lncRNA_mrna_promoters.gtf")

# lncRNA and mRNA gene bodies

lncrna_mrna_genebody <- rtracklayer::import("file_path/lncrna_mrna_genebody.gtf")

lncrna_mrna_genebody <- rtracklayer::import("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/11_consensus_peaks/gene_annotation_files/lncrna_mrna_genebody.gtf")


# Now  Let's also make a list of lncRNA gene_ids. Note the indexing.
# Double note the how handy it is to index to subsect a file.
lncrna_gene_ids <- lncrna_mrna_genebody$gene_id[lncrna_mrna_genebody$gene_type == "lncRNA"]

# same for mRNAs
mrna_gene_ids <- lncrna_mrna_genebody$gene_id[lncrna_mrna_genebody$gene_type == "protein_coding"]

```


# creating a data frame of peak overlaps.
Now we are going to start to make a really useful data-frame (df).
For each property/overlap we will make a new column. 

```{r creating an overlap features data.frame}

# Let's start with loading in the number of peaks each DBP has -- using length.

num_peaks_df <- data.frame("dbp" = names(filtered_consensus_peaks),
                           "num_peaks" = sapply(filtered_consensus_peaks, length))


# Now let's get the total amount of the genome covered by all the peaks for a given DBP.

num_peaks_df$total_peak_length <- sapply(filtered_consensus_peaks, function(x) sum(width(x)))

# Let's take a look at what we have so far (which DBP, has most peaks? most genome coverage?)
```

# expanding our num_peaks_df information
Great we are starting to build an informative DF to incorporate our results.
Now let's start looking at how many peaks overlap the genome feature files we made.

```{r adding more info into num_peaks_df}

# First let's get a count for each DBP that overlaps any promoter.
## This takes a minute to run :) Let's look count_peaks_per_feature function in meantime

promoter_peak_counts <- count_peaks_per_feature(lncrna_mrna_promoters, filtered_consensus_peaks, type = "counts")

# Take a look at "count_peaks_per_feature" maybe look at in debugging mode?

```

# extracting promoer_peak_counts data 
Above we made the object "promoter_peak_counts" now we can extract information from it

```{R xtracting promoer_peak_counts data }

# First let's get a count of how many promoter overlapped each DBP.
# we are making a new col and populating on the other side of assignment operator
# NOTE THIS TAKES ~10-15 min
num_peaks_df$peaks_overlapping_promoters <- rowSums(promoter_peak_counts)

# Now let's break these promoters into two groups "lncrna" and "mrna"
# We will use the gene_id objects we made above to index and separate them.

num_peaks_df$peaks_overlapping_lncrna_promoters <- rowSums(promoter_peak_counts[,lncrna_gene_ids])

# mrna promoter overlaps
num_peaks_df$peaks_overlapping_mrna_promoters <- 
  rowSums(promoter_peak_counts[,mrna_gene_ids])

# Let's take a look.

write_csv(num_peaks_df, "num_peaks_df.csv")

```

# extracting gene_body_peak_counts data 
Now let's do the same thing for gene-bodies
First we will make a gene-body object like we did for promoters above.

```{r extracting genebody overlpas counts data }

# Finding overlaps with gene_bodies (will take a few minutes again)
# Note this takes 10-15 min
genebody_peak_counts <- count_peaks_per_feature(lncrna_mrna_genebody, 
                                                filtered_consensus_peaks, 
                                                type = "counts")

# Now let's extract the overlaps the same way we did for promoters above

# All gene bodies
num_peaks_df$peaks_overlapping_genebody <- 
  rowSums(genebody_peak_counts)

# lncRNA gene bodies 
num_peaks_df$peaks_overlapping_lncrna_genebody <- rowSums(genebody_peak_counts[,lncrna_gene_ids])

# mRNA gene bodies
num_peaks_df$peaks_overlapping_mrna_genebody <- 
  rowSums(genebody_peak_counts[,mrna_gene_ids])

# let's take a look.

# Now let's write it out to a file -- will load this moving forward in class
# This is a time saver from having to regenerate the overlaps that took a bit ...

# NOTE: SET WORKING DIRECTORY TO 12_

# BE SURE YOU HAVE THE RIGHT FILE PATH to 12_
write_csv(num_peaks_df, "num_peaks_df.csv")

```

# Downloading TF annotation data set
We have a large group of different DBPs so let's import some annotaitons of which
ones are transcription factors etc. A paper in Cell is one of the better annotations --
however they are never perfect :)

Human TF annotations:
https://www.cell.com/cms/10.1016/j.cell.2018.01.029/attachment/ede37821-fd6f-41b7-9a0e-9d5410855ae6/mmc2.xlsx

Let's download the file using R (we could also use curl/wget in BASH)
```{r downloading TF annotation data}

url <- "https://www.cell.com/cms/10.1016/j.cell.2018.01.029/attachment/ede37821-fd6f-41b7-9a0e-9d5410855ae6/mmc2.xlsx"

destination_for_url <- "/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/TF_annotations.xlsx"


# to download we can use download.file

download.file(url, destination_for_url)

# The above is a handy bit of code for downloading data into R directly,but BASH works too

# Download the file locally and let's look it over to see what we might want.
```

# reading in DBP annotations as transcription factor
Now that we have the file we need to read it in. Since it's excel file we will use
a new 'read' function (we have used 'read.table' & 'read.csv' etc).

To this end we will use readX1 'read_excel' function.
```{R merging in TF annotation data}

human_tfs <- readxl::read_excel("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/TF_annotations.xlsx",
                                sheet = 2, skip = 1)


# let's rename the 4th column to indicate if it is a TF.
names(human_tfs)[4] <- "is_tf"


# now let's intersect gene names that are in our ChIP data and has TF identity.

length(which(tolower(num_peaks_df$dbp) %in% tolower(human_tfs$Name)))
```

# merging annotation file to num_peaks_df
Cool 437 are in there so let's merge the tables
```{R merging annotation file to num_peaks_df}

# first let's filter and grab the first 4 columns that match DBPs in num_peaks_df
human_tfs <- human_tfs[tolower(human_tfs$Name) %in% tolower(num_peaks_df$dbp), 1:4]


# adding new column names
names(human_tfs) <- c("ensembl_id",
                      "dbp",
                      "dbd",
                      "tf")

# Now comes the actual merge. Let's look at it first
# ?merge
# num_peaks_df <- read_csv("num_peaks_df.csv")

num_peaks_df <- merge(num_peaks_df, human_tfs, all.x = T)

# Let's check how many NAs -- we should have some missing values.
dim(num_peaks_df[is.na(num_peaks_df$tf),])

# Ok looks good let's write it out:
num_peaks_df <- num_peaks_df[,1:12]
write_csv(num_peaks_df, "file_path/num_peaks_df.csv")

# write_csv(num_peaks_df, "/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/num_peaks_df.csv")
```

# Occurence matrix of overlaps
Now let's also use our 'count_peaks_per_feature' function to use the 'occurence' parameter
We will make a matrix that records all the overlaps with all promoters.

```{r count_peaks_per_feature occurence matrix}

# note similar as above but using type = occurnece

promoter_peak_occurence <- count_peaks_per_feature(lncrna_mrna_promoters, filtered_consensus_peaks, 
                                               type = "occurrence")

# Let's double check that all lncrna & mrna genes are accounted for:

stopifnot(all(colnames(promoter_peak_occurence) == lncrna_mrna_promoters$gene_id))

# Great we will use this quite a bit moving forward so let's write it out! 
# we are going to write it out a tab seperated values (tsv) file
# moving forward we will use (?read.tsv) to read this in and save us all that run time :)
# DON'T OPEN THIS FILE IN THE ENVIRONMENT -- VERY LARGE MATRIX

write.table(promoter_peak_occurence, "/file_path/lncrna_mrna_promoter_peak_occurence_matrix.tsv")

# write.table(promoter_peak_occurence, "/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/lncrna_mrna_promoter_peak_occurence_matrix.tsv")

```

# Peak occurence as data.frame
Now let's make the peak occurence matrix into a DF for easier access -- but not computation.
```{R turnign peak_occurence matrix into data.frame}

# reading back in to be sure all is good.
# peak_occurence_df <- read.table("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/lncrna_mrna_promoter_peak_occurence_matrix.tsv")


# First make sure promoter_peak_occurence and lncrna_mrna_promoters are in the same order
stopifnot(all(colnames(promoter_peak_occurence) == lncrna_mrna_promoters$gene_id))
# cool all is good here even if read in above with read.table



# Now let's use the 'data.frame()' fucntion. Set up a bunch of colnames and populate them.

peak_occurence_df <- data.frame("gene_id" = colnames(promoter_peak_occurence),
                                "gene_name" = lncrna_mrna_promoters$gene_name,
                                "gene_type" = lncrna_mrna_promoters$gene_type,
                                "chr" = lncrna_mrna_promoters@seqnames,   
                                "3kb_up_tss_start" = lncrna_mrna_promoters@ranges@start,
                                "strand" = lncrna_mrna_promoters@strand,
                                "number_of_dbp" = colSums(promoter_peak_occurence))

# Go through each population index by highlighting and running


# Let's write out this data frame -- all this code & run time is now encapsulated in one .csv :)

write_csv(peak_occurence_df, "file_path/peak_occurence_dataframe.csv")

looking <- read_csv("peak_occurence_dataframe.csv")

# write_csv(peak_occurence_df, "/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/12_peak_features/peak_occurence_dataframe.csv")

# Cool feel free to open this one -- even in excel :) 
```

#############
ExeRcise
#############

Take a look at the peak_occurence_df and num_peaks_df -- find 3 things new that we haven't done yet. 

Simple solutions:

1) Which DBP has the most coverage of peaks in lncRNAs 
2) Which non TF has the most peaks and genome coverage. How does it compare to those that are not TFs
3) percentage of peaks that overlap promoters versus gene bodies

