---
title: "02_design_file"
author: "JR"
date: "8/8/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(tidyverse)

# install.packages("httr")
library(httr)

# install.packages("janitor")
library(janitor)

# install.packages("purrr")
library(purrr)

source("path/util/my_class_functions.R")

```

In the previous class we created a function that can go and get data from encode and 
make it an object in our environment "encode_file_info". What we would ideally like to do
is provide a list of experimenta_accessions from ENCODE and have encode info give us the data.

NOTE: using this approach we can make the design file from encode_file_info funciton. 
Thus making the design file for NF_CORE 100% reproducible in code !

FIRST, we need to learn a really awesome and powerful function called "MAP" in the "PURR" package.

# Introduction to map
Since one encode experimental accession has multiple files we can "MAP" 
all the files to a single entry for each experiment accession. This would 
be similar to putting a bunch of file paths in one excel "cell" 

Let's start with the basics.

# Using MAP
```{r MAP example}

# In this example we'll take the digits 1 through 10 and "map" each to the rnorm function. 
map_example <- 1:10 %>%
  map(rnorm, n = 1000) 

# It returns a nested list, let's look:
summary(map_example)

# Let's look with table
table(summary(map_example))

# let's index and look
map_example[[1]]

hist(map_example[[1]])

# another one
map_example[[2]]

# indexing to the first component of map vector
map_example[[1]][[1]]

```

# MAP encode_file_info function output

We can use map to call this encode_file_info function for each experiment accession
and return each experimental_accession as an entry in a data.frame 
Specifically one row associated with experiment acession (ENCSR) will have all the encode files (ENCFF)
So one row will have many files in it -- similar to have a bunch of files in a single excel "cell"

First let's read in the RNAseq samples we downloaded from ENCODE.

```{r reading in sample sheet}

# We'll also rename this Accession column to clarify between experiment_accession and file_accession.
samples <- read.table("data/samples.txt",
                      sep = "\t", skip = 1, header = T) %>%
  dplyr::rename(experiment_accession = Accession) 

# It's seems mundane but starting here is the best way to make a "reproducible" sample sheet.
# Bottom line: the download to code to analysis is the way to reproducibility (this worked a year later :)
```

# Mapping each experiment to the experiment_accession

Each experiment accession has multiple files associated with the accession
Let's use map to map each experiment to it's experiment accession and clean up sample sheet

```{r MAP to experiment_accession}

?mutate
?map

# We are making a new column in samples called file_info. This is created
# by mapping experiment_accession and putting the variable through encode_file_info function.
# This will map the information (8 fields) retrieved by encode_file_info function by 
# experiment_accession in a column called file_info.

samples <- samples %>%
  # note function inside a function being called.
  # the "experiment_acession" col will be read into encode_file_info function
  # ~ means pass the experiment_accession value into encode_file_info
  # (.x) is a generic symbol meaning pass what is in memory == experiment_acession
  mutate(file_info = map(experiment_accession, ~ encode_file_info(.x)))

# Thus for each accession in experiment_accession mutate will use the function MAP
# to map each experimental file to the "file_info" column and keep the data there.

```

# Unnesting the data from MAP
This is a bit hard to read in this format, so we can unnest the data.frames in the file_info column using the unnest command.
This will duplicate the remaining columns to match the number of lines in the nested data.frame.

```{r unnesting MAP}

?unnest
# This function will extract the mapped info and make new cols and rows
# We also need to tell it which column we want to unnest.

# note samples is currently 5 rows
samples <- samples %>%
  unnest(cols = file_info)
# Now samples is 20 rows from unnesting the 5 previous rows.
# This is because each experimental accession had 4 files for 5 exp accession = 20 rows

```

# cleaning up sample sheet
Now we have all the information we neeed in this data.frame and we just need to clean it up.

```{r sample sheet curration: adding replicate number for each RNAseq experiment}
# Let's number our replicates and create a new column called sample id 
# where we join the experiment accesion and the rep number

samples <- samples %>%
  group_by(experiment_accession) %>%
  mutate(rep_number = as.numeric(factor(replicate))) %>%
  unite(sample_id, experiment_accession, rep_number, sep = "_rep", remove = F)

 # unite will make a new column in samples (piped in) called sample_id
  # here we will combine experiment accession and rep number with _rep in between
  # This is a handy handle for the data to access later
```

# getting rid of non essential data
Now let's get rid of all that data we don't need! 

```{r removing data from sample sheet}

# We're just selecting a subset of columns from samples with dplyr::select
samples <- samples %>%
   dplyr::select(accession, sample_id, experiment_accession, Assay.title,
                Biosample.summary, md5sum, paired_end_identifier) %>%
  
  # here we are going to start to build the fastq file name for NF_core input
  unite(fastq_file, sample_id, paired_end_identifier, sep = "_read", remove = F)

# Now we have a new col fastq_file with the information for each fastq file
# We will keep adding to this until we get to the final NF_Core input file name.
```

# Setting up filename for NF_CORE rnaseq pipeline
Now let's make the full filename for the fastq files. 
For the nf-core/rnaseq pipeline, the paired-end reads need to be named 
with the read number in the filename. 

```{r filename for NF_CORE rnaseq pipeline: adding .fastq.gz to fastq_file}

# We are now making a new column with .fastq.gz extension and then unite to make fastq file name.

samples <- samples %>%
  mutate(fq_extension = ".fastq.gz") %>%
  unite(fastq_file, fastq_file, fq_extension, sep = "", remove = F) %>%
  
  # This original file column will be used along with the new name column to rename the fastq files.
  unite(original_file, accession, fq_extension, sep = "")

# nice we see a fastq file column and can be used as input into NF_CORE
```

# all code above compiled in one chunk :)

We broke this down into parts, so you can understand what is happening, 
but just note that you can write all of this in one block and read it like a sentence.

```{r compiled code to get samplesheet -- all code above compiled into 1 chunk}

# This is simply all the code above in one chunk.

# CAN SKIP IF TIME IS LOW.

# samples <- read.table("samples.txt",
#                       sep = "\t", skip = 1, header = T) %>%
#   dplyr::rename(experiment_accession = Accession) %>%
#   mutate(file_info = map(experiment_accession, ~ encode_file_info(.x))) %>%
#   unnest(file_info) %>% 
#   group_by(experiment_accession) %>%
#   mutate(rep_number = as.numeric(factor(replicate))) %>%
#   unite(sample_id, experiment_accession, rep_number, sep = "_rep") %>%
#   dplyr::select(sample_id, accession, Assay.title,
#                 Biosample.summary, md5sum, paired_end_identifier) %>%
#   unite(fastq_file, sample_id, paired_end_identifier, sep = "_read", remove = F) %>%
#   mutate(fq_extension = ".fastq.gz") %>%
#   unite(fastq_file, fastq_file, fq_extension, sep = "", remove = F) %>%
#   unite(original_file, accession, fq_extension, sep = "")


```

# renaming fastq files to fit sample sheet
This cleaned up version of the samplesheet is good to go!
Now we want to rename the fastq files to the fastq name we just made.

```{r rename fastq files to samplesheet id}
# Rename the fastq files so that they contain the sample ID.
rename_script <- samples %>%
  # removes grouping in samples
  ungroup() %>%
  dplyr::select(fastq_file, original_file) %>%
  mutate(command = "mv") %>%
  unite(command, command, original_file, fastq_file, sep = " ")
# The result of this is that each row is a bash command.

# We can write this out as a bash script with ?write_lines 
# We include a shebang header line so that the script is interpreted by bash.
?write_lines

write_lines(c("#!/bin/bash", rename_script$command), "rename.sh")

# Now cd fastq and "chmod u+x rename.sh
# then ./rename.sh

# >>> voila all the files are renamed
```

# md5sum check on file downloads
Additionally from all of this information we've gathered we can create a text file to run the md5sum check.

```{r md5sum of fastq}
# Let's create an md5.txt to run the checksums
# We will use the same approach for the mv script.

md5 <- samples %>% 
  ungroup() %>%
  dplyr::select(md5sum, fastq_file) %>%
  # note double space for md5 requirements 
  unite(line, md5sum, fastq_file, sep = "  ")

# writting out md5$line col that has the command in it from above.
write_lines(md5$line, "md5.txt")

# Now let's run it from R to BASH with "system()" function
 md5_results <- system("cd fastq; md5sum -c md5.txt")
 
# md5 -c md5.txt > md5_check.txt
 
 #TODO here is a solution to run system and get object:
 # To output the md5_check as an R vector, add the "intern = TRUE" argument to the system() function:

# md5_results <- system("cd 17/fastq; md5sum -c md5.txt", intern = TRUE)

 #TODO check on system runs populating variables
# Nice they all look correct -- but let's save this result
# write.csv(md5_results, "md5_check.csv")
```

# finalizing sample sheet for NF_CORE
Finally, we can write out a nicely formatted sample sheet 
that we will use downstream for further analysis of the read counts in R.

```{r Reorganizing to make a sample sheet for DEseq2}
# Let's create the sample sheet that we will use later
# to do the RNA-seq analysis in R.

samples <- samples %>%
  dplyr::rename(fastq = fastq_file,
                seq_type = Assay.title,
                sample_name = Biosample.summary) %>%
  # The minus sign will remove this column -- which we no longer need.
  dplyr::select(-original_file) 

```

# final organization of sample sheet

Now that we have it cleaned up, let's create one line for each replicate
where the fastq read 1 and read 2 are in the same row.
```{R organizing samplesheet }


# For this we will use the pivot wider function
# We need to tell the pivot_wider function which unique column combinations will specify each new row. 

# PIVOT WIDER by "paired_end_identifier" 
samplesheet <- samples %>%
  #id_cols is a parameter in pivot wider to select the cols
  # the paired end identifier becomes the "marienette" string of the data-frame.
  # There are two values and thus all the current cols will be split into 2 (one for each pe-id)
  pivot_wider(id_cols = c("sample_id", "seq_type", "sample_name"),
              names_from = paired_end_identifier,
              values_from = c("fastq", "md5sum"))


# Cleaning up sample sheet (removing spaces - re-arrange etc)
samplesheet <- samplesheet %>%
  
# cleaning up column "sample_name" that has spaces in it to replace with underscore
mutate(condition = gsub(" ", "_", sample_name) %>% tolower()) %>%
  
# splitting up "sample_id" to extract replicate number (by "_" )
separate(sample_id, into = c("experiment_accession", "replicate"), 
           remove = FALSE, sep = "_") %>%
  
# replicate col values came from sample id and are currently rep1 or rep2
# we want to remove the "rep" with gsub to "R" and iterative using mutate
mutate(replicate = gsub("rep", "R", replicate)) %>%
  
# we are writting over the sample_name col and uniting condition and replicate info 
# into the previous sample_name col. syntax: (data frame - implied from tidy, new_col_name, what to unite)
unite(sample_name, condition, replicate, sep = "_", remove = FALSE)

# here we are just changing the hepg2 total name to hepg2_total
#TODO leaving out and could fix matching condition but not needed
# samplesheet$condition[samplesheet$condition == "hepg2"] <- "hepg2_total"


# FINAL cleanup of col names etc.
samplesheet <- samplesheet %>%
  
  mutate(cell_type = "hepg2",
         condition = gsub("hepg2_", "", condition)) %>%
  
  dplyr::select(sample_id, sample_name, replicate, condition,
                cell_type, seq_type, fastq_1, fastq_2, md5sum_1,
                md5sum_2)

# that was a lot of work so let's save for future use :) 
# Writting this out to 17_ test_2
write_csv(samplesheet, "samplesheet.csv")
```

# design file
We will eventually need a design file for RNA-seq (DEseq2)
We are really close with samples sheet so let's finalize that.

# Future RNAseq NF_Core runs require design file
Newwer versions of NF_Core require a design file the version we are running does not.

The design file needs to be in a specific format (as we saw with nf-core/chipseq)
It needs the following columns:

# sample,fastq_1,fastq_2,strandedness

Let's create a sample column using mutate -- and we'll clean up the names

```{r creating design file}

base_path <- "/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/17_API_RNASEQ/fastq/"

# base_path <- paste0("/scratch/Shares/rinnclass/CLASS_2022/JR/CLASS_2022/class_exeRcises/analysis/17_API_RNASEQ/fastq/")

# There's spaces in the names currently and we need to get rid of those.
design <- samplesheet %>%
  
# we will modify 4 cols adding "sample" and "strandedness"
# We will also add a file path to "fastq_1" and _2 cols
# first changing "sample_name" and adding strandedness col
mutate(sample = gsub(" ", "_", sample_name) %>% tolower(),
         strandedness = "unstranded",
       
# now pasting file path into fastq_1 and _2 cols
#TODO MAKE SURE YOU HAVE THE RIGHT FILE PATH TO YOUR FASTQ FILES

# base_path <- "/path to your fastq" e.g. "/scratch/Shares/rinnclass/CLASS_2022 ....."
         fastq_1 = paste0(base_path, fastq_1),
         fastq_2 = paste0(base_path, fastq_2)) %>%
  
  
  # We aim to retrieve replicate number for sample_id and make it just a number 1 or 2 
  separate(sample_id, into = c("experiment_accession", "replicate"), sep = "_", remove = FALSE) %>%
  
  # getting rid of the rep in front of rep1 and rep2
  mutate(replicate = gsub("rep", "", replicate)) %>%
  
  # Now we gather just the columns we need
  dplyr::select(sample, fastq_1, fastq_2, strandedness)


# getting rid of _r1 and _r2 from sample col
# note we know which replicate is which from fastq col
design <- design %>%
  mutate(sample = gsub("_r1", "", sample),
         sample = gsub("_r2", "", sample))

# let's test that the path to fastq works:
all(sapply(design$fastq_1, file.exists))
all(sapply(design$fastq_2, file.exists))

# Now we can write this out for input into nextflow

# Writting out to 17_ test_2
write_csv(design, "design.csv")

write_csv(samples, "samples.csv")

write_csv(samplesheet, "samplesheet.csv")
```


Now we have the raw files downloaded and the samplesheet needed for downstream analyses (ex, differential expression). So the next step is to run nf-core/rnaseq on these samples.
We'll do this in the RNAseq II.

