---
title: "02_plotting_consensus_peaks"
author: "JR"
date: "8/10/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
library(ggplot2)
source("../../util/my_class_functions.R")
source("../../util/_setup.R")

```

# loading in peak_features env objects from 01_create_consensus_peaks
```{r laod env objects}

load("../01_create_consensus_peaks/results/peak_features.RData", verbose = T)

```


# Peaks per dbp
```{r plotting peak features}

# First let's look at a histogram of peak#/DBP

 ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_histogram(bins = 70)
# saving
ggsave("figures/num_peaks_hist.pdf")

```


# plotting num_peaks versus genome coverage.
```{r peaks vs coverage}


ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length)) +
  geom_point() + 
  geom_smooth(method = "gam", se = TRUE, color = "black", lty = 2)+
  ylab("BP covered") +
  xlab("Number of peaks") +
  ggtitle("Peak count vs. total bases covered")

# saving 
ggsave("figures/peak_num_vs_coverage.pdf")


```
Result: there is a linear relationship between number of peaks and total coverage


# plotting num peaks on promoters

```{r number of DBPS on promoters}


ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_promoters)) +
  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping promoters") +
  ggtitle("Relationship Between Number of DBP Peaks and Promoter Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=FALSE, formula = 'y ~ x',
              color = "#a8404c") +
  stat_regline_equation(label.x = 35000, label.y = 18000) +
  ylim(0,60100) +
  xlim(0,60100)

ggsave("figures/peak_num_vs_promoter_coverage.pdf")

# w
# peaks you stop increaseing binding to promoters.
# maybe it leaks into the gene body let's check
```
Result: saturation of binding events -- as you get more peaks 
you stop increasing binding to promoters -- probably saturated.

# peak Coverage on gene bodies
```{r peak coverage on gene bodies}

ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_genebody)) +
  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping genes") +
  ggtitle("Relationship Between Number of DBP Peaks and Gene Body Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=F, formula = 'y ~ x',
              color = "#a8404c") +
  stat_regline_equation(label.x = 35000, label.y = 18000) +
  ylim(0,60100) +
  xlim(0,60100)

# saving
ggsave("figures/4_peak_num_vs_gene_body_coverage.pdf")
```
Result: Gene bodies explain almost all the places of binding in the genome




# Density plot of binding events
Let's make a density plot of num DBPs bound per promoter

```{r density plot of DBP localization events}

ggplot(peak_occurence_df, aes(x = number_of_dbp)) +
geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  theme_paperwhite() +
  xlab(expression("Number of DBPs")) +
  ylab(expression("Density")) +
  ggtitle("Promoter binding events",
          subtitle = "mRNA and lncRNA genes") 

# saving
ggsave("figures/num_binding_events_per_promoter.pdf")

```
Result: very interesting that the promoter binding is bimodal !
Result: most promoters have upto 100 dbps then a lag and 2nd dist at ~200dbs
# RESULT: There are two types of promoter binding - (1) normal (2) super-binders


# promoters with out binding events
Lets find how many promoters don't have any DBPs bound
```{r prmoters with out binding events}

unbound_promoters <- peak_occurence_df %>% 
  filter(peak_occurence_df$number_of_dbp < 1)

# how many are there?
nrow(unbound_promoters)
# so there are only a few 6,720 promoters that don't have binding evetns (~10%)

#  let's put it in a folder called results. We will always use this folder structure
write_csv(unbound_promoters, "results/unbound_promoters.csv")

```

# lncRNA versus mRNA promoter binding
Let's compare the binding patterns of lncRNA vs mRNA promoters.
```{r lncrna vs mrna promoter binding}

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_point(aes(y = peaks_overlapping_lncrna_promoters), color = "red") +
  geom_point(aes(y = peaks_overlapping_mrna_promoters), color = "black") +
  stat_regline_equation(aes(y = peaks_overlapping_lncrna_promoters), color = "red") +
  stat_regline_equation(aes(y = peaks_overlapping_mrna_promoters), color = "black", label.y = 20000) +
  geom_smooth(aes(y = peaks_overlapping_lncrna_promoters), method = "lm", se = FALSE, formula = "y ~ x") +
  geom_smooth(aes(y = peaks_overlapping_mrna_promoters), method = "lm", se = FALSE, formula = "y ~ x")

# saving
ggsave("figures/lncRNA-mRNA-promoter_binding.pdf")



# This is just a fancier version of the same thing -- note mutate to clean up names
num_peaks_dfl <- num_peaks_df %>%
  dplyr::select(-peaks_overlapping_promoters) %>%
  pivot_longer(cols = peaks_overlapping_lncrna_promoters:peaks_overlapping_mrna_promoters,
               names_to = "gene_type",
               values_to = "peaks_overlapping_promoters") %>%
  mutate(gene_type = gsub("peaks_overlapping_", "", gene_type))

# plotting
ggplot(num_peaks_dfl, aes(x = num_peaks, y = peaks_overlapping_promoters, 
                         col = gene_type)) +
         geom_point() +
         geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se = FALSE, formula = "y ~ x") +
  stat_regline_equation() +
  scale_color_manual(values = c("#a8404c", "#424242"))+
  xlab("Peaks per DBP") +
  ylab("Peaks Overlapping Promoters") +
  ggtitle("Number of DBP Peaks and Promoter Overlaps")

# saving
ggsave("figures/peaks_overlaps_relationship_by_gene_type.pdf", height = 5, width = 8)


```




